### 2020-09-09 1.P259.pDC_data.cleaning.Rmd
init public manuscript repo

---
title: 'P259: Combined data cleaning'
subtitle: "Dendritic cells (pDC)"
author: "Kim Dill-McFarland, kadm@uw.edu"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
always_allow_html: true
---
# Background

P259.1: pDC were isolated from 4 healthy donors' whole blood or buffy coat. Cells were adapted to media with IL3 and then cultured with eosinophil (EOS) supernatant and/or human rhinovirus (RV).

P259.2: pDC were isolated from 7 asthmatic donors' whole blood, 4 of which were on Anti-IL5 therapy. Cells were adapted to media with IL3 and then cultured with or without RV.

The purpose of this workflow is to complete quality control and data cleaning of metadata and RNA-seq libraries generated from the above experiments. This includes 1) removing low coverage libraries, 2) filtering rare genes, 3) removing outlying libraries, and 4) normalizing for RNA composition.

# Setup
Load packages

```{r message=FALSE, warning=FALSE}
# Data manipulation and figures
library(tidyverse)
  #Multi-panel figures
  library(cowplot)
  # Modify ggplot figures to non-overlapping text labels
  library(ggrepel)
  # Modify ggplot data order within facets
  library(drlib)
  # Plot log scales
  library(scales)
# Reading Excel files
library(readxl)
# Working with dates
library(lubridate)

# Empirical analysis of digital gene expression data
## Data normalization
library(edgeR)

# Print pretty table to knit file
library(knitr)
library(kableExtra)
  options(knitr.kable.NA = '')

#Define colors for plots
group.cols <- c("none:none"="#dadaeb",
               "none:AntiIL5"="#9e9ac8",
               "none:EOS.supp"="#54278f",
               "HRV:none"="#c7e9c0",
               "HRV:AntiIL5"="#74c476",
               "HRV:EOS.supp"="#006d2c",
               "flu:none"="#fdae6b",
               "flu:AntiIL5"="#e6550d")
samp.cols <- c("AC1"="#969696",
               "AC2"="#a6cee3",
               "AC3"="#1f78b4",
               "AC4"="#b2df8a",
               "AC5"="#33a02c",
               "AT1"="#fb9a99",
               "AT2"="#e31a1c",               
               "AT3"="#fdbf6f",
               "AT4"="#ff7f00",
               "donor1"="#cab2d6",
               "donor2"="#6a3d9a",
               "donor3"="#ffff99",
               "donor4"="#b15928")
```

Set seed

```{r}
set.seed(4389)
```

# Read in and format data 
## Metadata

```{r echo=FALSE, message=FALSE}
meta <- read_csv("data_clean/P259_pDC_metadata.csv") %>% 
  #Set factosr and orders
  mutate(IL5 = factor(IL5, levels = c("none","AntiIL5","EOS.supp")),
         virus = factor(virus, levels=c("none","HRV")),
         virus.detail = factor(virus.detail,
                               levels=c("none","oldHRV","newHRV"))) %>% 
  arrange(libID)
```

## Counts table

Read in counts table from P259.1 and P259.2 (`data_raw/*combined_counts.csv`) and then combine to one data frame.

```{r echo=FALSE, message=FALSE}
count.raw1 <- read_csv("data_raw/P216-1_CCLKLANXX_180806_combined_counts.csv") %>% 
  # Remove '_CE03HANXX' suffix from variable names
  rename_at(.vars = vars(ends_with("_CCLKLANXX")),
            .funs = ~sub("_CCLKLANXX", "", .))
```

```{r echo=FALSE, message=FALSE}
count.raw2 <- read_csv("data_raw/P259-2_CE03HANXX_191107_combined_counts.csv") %>% 
  # Remove '_CE03HANXX' suffix from variable names
  rename_at(.vars = vars(ends_with("_CE03HANXX")),
            .funs = ~sub("_CE03HANXX", "", .))
```

```{r echo=FALSE, message=FALSE}
count.all <- full_join(count.raw1, count.raw2,
                       by = "geneName")
```

Read in and filter associated gene key (`data_raw/2020.06.18_HGNC.gene.key.txt`, downloaded from <https://www.genenames.org/download/custom/>) to protein coding (pc) genes that occur in the count data set and have valid hgnc symbols.

```{r echo=FALSE, message=FALSE}
key <- read_tsv("data_raw/2020.06.18_HGNC.gene.key.txt",
                na=c("")) %>% 
  # rename variable to match count data
  dplyr::rename(geneName = "Ensembl gene ID",
                hgnc_symbol = "Approved symbol",
                gene_biotype = "Locus group") %>% 
  #remove genes w/o ensembl ID
  drop_na(geneName) %>% 
  #collapse multi-annotations
  group_by(geneName) %>% 
  mutate(across(c(gene_biotype, hgnc_symbol), 
                ~paste(unique(.), collapse=", "))) %>% 
  distinct()

key.pc <- key %>% 
  # Keep protein coding genes only
  filter(gene_biotype == "protein-coding gene") %>% 
  # Keep genes in counts data
  filter(geneName %in% count.all$geneName) %>% 
  # Remove duplicate entries
  distinct(geneName, .keep_all=TRUE) %>% 
  arrange(geneName)
```

Then filter the count data to pc genes and save.

```{r echo=FALSE, message=FALSE}
count.pDC <- count.all %>% 
  filter(geneName %in% key.pc$geneName) %>% 
  arrange(geneName) %>% 
  #Order columns and keep only pDC data
  select(geneName, all_of(meta$libID))
```

```{r}
#Save counts table
write_csv(count.pDC, "data_clean/P259_pDC_counts.csv")
```

```{r echo=FALSE}
#Remove intermed files
rm(count.raw1, count.raw2, count.all, key)
```

#### Verify data order

To ensure that all data match, check that all libraries in the count data are present and in the same order in the metadata. And vice versa.

```{r echo=FALSE, message=FALSE}
identical(meta$libID, colnames(count.pDC)[-1])
```

## Sample summary

At this point, the following objects exist in our environment. Protein coding = pc

```{r echo=FALSE}
ls()
```

And the sample numbers are

```{r echo=FALSE}
meta %>% 
  group_by(experiment, asthma,IL5,virus) %>% 
  tally() %>% 
  arrange(experiment, asthma,IL5,virus) %>%

  kable(align="c") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
  collapse_rows(1:2, valign="top")
```

# Data cleaning
## Median CV coverage vs. mapped duplicate reads

Compare the median coefficient of variation (CV) coverage (`median_cv_coverage`) and percent alignment of reads with duplicates (`mapped_reads_w_dups`).

```{r echo=FALSE, fig.height=4, fig.width=8.5}
# Plot libraries
meta %>% 
  #Create color group for total sequences
  mutate(col.group = ifelse(total_sequences <= 1e5, "<100,000",
                     ifelse(total_sequences <= 3e5, "<300,000",
                     ifelse(total_sequences <= 5e5, "<500,000",
                            ">500,000")))) %>%

ggplot(aes(median_cv_coverage, mapped_reads_w_dups)) +
  geom_point(aes(color=col.group, shape=experiment), size=2) +
  #Label points outside cutoffs
  geom_text_repel(data=filter(meta, median_cv_coverage > 0.6 | 
                              mapped_reads_w_dups < 0.7),
            aes(label=libID), show.legend = FALSE) +
 #Beautify
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x="Median CV coverage", y="Mapped reads with duplicates",
       color="Total sequences") +
  lims(x=c(0,1), y=c(0,1)) +
  scale_color_manual(values = c("#cbc9e2","#9e9ac8","#756bb1","#54278f"))
```

## Total aligned counts

Plot total counts per library. Here, we compare our standard cutoff of 500,000 to a less strict cutoff of 100,000, because `r nrow(filter(meta, total_sequences <= 5E5))` pDC libraries would be removed at the standard cutoff.

```{r echo=FALSE, fig.height=4, fig.width=8.5}
meta %>% 
  #Create color group for total sequences
  mutate(col.group = ifelse(total_sequences <= 1e5, "<100,000",
                     ifelse(total_sequences <= 3e5, "<300,000",
                     ifelse(total_sequences <= 5e5, "<500,000",
                            ">500,000")))) %>%

ggplot(aes(x=reorder_within(libID, by=total_sequences,
                            within=experiment), 
           y=total_sequences, fill=col.group))  +
  geom_col() +
  # Facet by cell type
  facet_grid(~experiment, scales="free_x", space="free") +
  # Add cutoff line
  geom_hline(yintercept = 1E5) +
  geom_hline(yintercept = 5E5, lty=2) +
  # Beautify
  theme_classic() +
  theme(axis.text.x = element_text(size=rel(0.75),
                                   angle = 90, hjust = 1),
        legend.position = "bottom") +
  labs(x="Library", y="Total aligned counts\n(Log scale)", fill="") +
  scale_x_reordered() +
  scale_fill_manual(values = c("#cbc9e2","#9e9ac8","#756bb1","#54278f")) +
  scale_y_continuous(trans = 'log10',
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x)))
```

## Filter by quality and coverage

Thus, the following libraries with questionable metrics (highlighted in red in HTML format) may be removed.

```{r echo=FALSE}
to_remove <- meta %>% 
  filter(total_sequences <= 5E5 | 
         median_cv_coverage > 0.6 |
         mapped_reads_w_dups < 0.75)

to_remove %>% 
  mutate(median_cv_coverage = cell_spec(median_cv_coverage, "html",
                              color = ifelse(median_cv_coverage > 0.6,
                                             "red", "darkgrey")),
         mapped_reads_w_dups = cell_spec(mapped_reads_w_dups, "html",
                              color = ifelse(mapped_reads_w_dups < 0.75,
                                             "red", "darkgrey")),
         total_sequences = cell_spec(total_sequences, "html",
                              color = ifelse(total_sequences <= 1E5,
                                             "red", "darkgrey"))) %>%
  select(experiment, libID, donorID, asthma,IL5,virus,
      median_cv_coverage, mapped_reads_w_dups, total_sequences) %>% 
  arrange(libID) %>%

  kable(align="c", format = "html", escape = FALSE) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

First, 3 libraries with < 100,000 sequences are removed. Since this removes the media (none:none) sample for donor AC1, all other AC1 samples are also removed. Of note, if the other AC1 samples are retained, they appears as PCA outliers in the next step.

```{r}
#Metadata
meta.filter <- meta %>% 
  # CV coverage and alignment
  filter(total_sequences >= 1E5) %>% 
  filter(donorID != "AC1") %>% 
  droplevels()

#Count data
count.pDC.filter <- count.pDC %>% 
  # Keep libraries (columns) remaining filtered metadata table (rows)
  dplyr::select(geneName, as.character(meta.filter$libID))
```

## Filter PCA outliers

Next, samples are assessed in PCA to determine if remaining questionable quality samples (labeled with libID) appear as PCA outliers.

```{r echo=FALSE, warning=FALSE, message=FALSE}
PCA.all <- count.pDC.filter %>% 
  column_to_rownames("geneName") %>% 
  #Convert to log counts per million
  cpm(., log=TRUE) %>% 
  t() %>% 
  #Calc PCA
  prcomp()

#Extract axes labels with percent contribution of PC
PC1.label <- paste("PC1 (", summary(PCA.all)$importance[2,1]*100, "%)", sep="")
PC2.label <- paste("PC2 (", summary(PCA.all)$importance[2,2]*100, "%)", sep="")

# Extract PC values
PCA.dat <- as.data.frame(PCA.all$x) %>% 
    rownames_to_column("libID") %>%
    # Select PCs for plotting
    dplyr::select(libID, PC1:PC3) %>% 
    # Merge with metadata
    left_join(meta.filter, by="libID")

PCA1 <- ggplot(PCA.dat, aes(PC1, PC2, color=virus:IL5)) +
      geom_point(size=2, stroke=1.5) +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nlogCPM") +
      coord_fixed(ratio=1) +
      scale_shape_manual(values=c(21,16)) +
      guides(color=guide_legend(title.position="top", 
                                title.hjust = 0.5))+
      scale_color_manual(values = group.cols)

PCA2 <- ggplot(PCA.dat, aes(PC1, PC2, color=donorID)) +
      geom_point(size=2, stroke=1.5) +
      geom_text_repel(data=filter(PCA.dat, 
                                  libID %in% to_remove$libID),
                      aes(label=libID), show.legend = FALSE) +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nlogCPM") +
      coord_fixed(ratio=1) +
      scale_color_manual(values=samp.cols)
```

```{r PCA.outlier, echo=FALSE, message=FALSE}
PCA.all.sd <- PCA.dat %>% 
  group_by(virus, IL5) %>% 
  #Calculate PC mean std deviation
  summarise(.groups="keep",
    PC1.mean = mean(PC1),
    PC1.sd = sd(PC1),
    PC2.mean = mean(PC2),
    PC2.sd = sd(PC2)) %>% 
  #Calculate +/- 3 sd limits
  mutate(
    PC1.min = PC1.mean-(3*PC1.sd),
    PC1.max = PC1.mean+(3*PC1.sd),
    PC2.min = PC2.mean-(3*PC2.sd),
    PC2.max = PC2.mean+(3*PC2.sd))

PCA.all.dat <- PCA.dat %>%   
  full_join(PCA.all.sd) %>% 
  #ID potential outliers
  mutate(col.group = ifelse(PC1 > PC1.max | PC1 < PC1.min |
                            PC2 > PC2.max | PC2 < PC2.min, 
                            "potential outlier", "okay"))

PCA3 <- PCA.all.dat %>%       
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color=col.group),size=2) + 
  geom_text_repel(data=filter(PCA.all.dat,
                            col.group == "potential outlier"),
            aes(label=libID), show.legend = FALSE) +
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, 
       title="pDC experiment\nlogCPM") +
  coord_fixed(ratio=1) +
  scale_color_manual(values = c("#969696","#b10026"))
```

```{r echo=FALSE, fig.height=8, fig.width=8.5}
plot_grid(PCA1,PCA2,PCA3, align = "v", ncol=1, axis="l")
```

Donor AC2 appears to be an outlier for all of its samples and will be removed.

```{r}
#Metadata
meta.filter2 <- meta.filter %>% 
  filter(donorID != "AC2") %>% 
  droplevels()

#Count data
count.pDC.filter2 <- count.pDC.filter %>% 
  # Keep libraries (columns) remaining filtered metadata table (rows)
  dplyr::select(geneName, as.character(meta.filter2$libID))
```

Re-assess PCA.

```{r echo=FALSE, warning=FALSE, message=FALSE}
PCA.all <- count.pDC.filter2 %>% 
  column_to_rownames("geneName") %>% 
  #Convert to log counts per million
  cpm(., log=TRUE) %>% 
  t() %>% 
  #Calc PCA
  prcomp()

#Extract axes labels with percent contribution of PC
PC1.label <- paste("PC1 (", summary(PCA.all)$importance[2,1]*100, "%)", sep="")
PC2.label <- paste("PC2 (", summary(PCA.all)$importance[2,2]*100, "%)", sep="")

# Extract PC values
PCA.dat <- as.data.frame(PCA.all$x) %>% 
    rownames_to_column("libID") %>%
    # Select PCs for plotting
    dplyr::select(libID, PC1:PC3) %>% 
    # Merge with metadata
    left_join(meta.filter2, by="libID")

PCA1 <- ggplot(PCA.dat, aes(PC1, PC2, color=virus:IL5)) +
      geom_point(size=2, stroke=1.5) +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nlogCPM") +
      coord_fixed(ratio=1) +
      scale_shape_manual(values=c(21,16)) +
      guides(color=guide_legend(title.position="top", 
                                title.hjust = 0.5))+
      scale_color_manual(values = group.cols)

PCA2 <- ggplot(PCA.dat, aes(PC1, PC2, color=donorID)) +
      geom_point(size=2, stroke=1.5) +
      geom_text_repel(data=filter(PCA.dat, 
                                  libID %in% to_remove$libID),
                      aes(label=libID), show.legend = FALSE) +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nlogCPM") +
      coord_fixed(ratio=1) +
      scale_color_manual(values=samp.cols)
```

```{r PCA.outlier2, echo=FALSE, message=FALSE}
PCA.all.sd <- PCA.dat %>% 
  group_by(virus, IL5) %>% 
  #Calculate PC mean std deviation
  summarise(.groups="keep",
    PC1.mean = mean(PC1),
    PC1.sd = sd(PC1),
    PC2.mean = mean(PC2),
    PC2.sd = sd(PC2)) %>% 
  #Calculate +/- 3 sd limits
  mutate(
    PC1.min = PC1.mean-(3*PC1.sd),
    PC1.max = PC1.mean+(3*PC1.sd),
    PC2.min = PC2.mean-(3*PC2.sd),
    PC2.max = PC2.mean+(3*PC2.sd))

PCA.all.dat <- PCA.dat %>%   
  full_join(PCA.all.sd) %>% 
  #ID potential outliers
  mutate(col.group = ifelse(PC1 > PC1.max | PC1 < PC1.min |
                            PC2 > PC2.max | PC2 < PC2.min, 
                            "potential outlier", "okay"))

PCA3 <- PCA.all.dat %>%       
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color=col.group),size=2) + 
  geom_text_repel(data=filter(PCA.all.dat,
                            col.group == "potential outlier"),
            aes(label=libID), show.legend = FALSE) +
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, 
       title="pDC experiment\nlogCPM") +
  coord_fixed(ratio=1) +
  scale_color_manual(values = c("#969696","#b10026"))
```

```{r echo=FALSE, fig.height=10, fig.width=8.5}
plot_grid(PCA1,PCA2,PCA3, align = "hv", ncol=2, axis="l")
```

Additionally, lib44585 was flagged for poor-quality above and is labeled in the PCA. While this sample is somewhat of an outlier in PCA, it is not > 3 s.d. away from its group's mean. Thus, it will remain and be re-assessed after gene filtering.

Therefore, the filtered data have

* max median CV coverage = `r round(max(meta.filter2$median_cv_coverage), digits=2)`
* min mapped reads with duplicates = `r round(min(meta.filter2$mapped_reads_w_dups), digits=2)`
* min total sequences = `r scales::comma_format()(min(meta.filter2$total_sequences))`

## Normalize for RNA composition

Create DGEList object

```{r}
dat.pDC.filter2 <- DGEList(
  #count table. move gene names to column names
  counts=as.matrix(column_to_rownames(count.pDC.filter2, "geneName")),
  #metadata
  samples=meta.filter2,
  #keep genes in count table
  genes=filter(key.pc, geneName %in% count.pDC.filter2$geneName))
```

## Filter rare genes

The raw gene sets contain highly variable, low abundance/rare genes. Genes must be at least 1 CPM in at least 3 samples to remain.

```{r}
source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/RNAseq_rare_gene_filter.R")

rare.gene.filter(dat.pDC.filter2, 
                 min.CPM=1,
                 min.sample=3, 
                 name="dat.pDC.filter2.abund")
```

```{r echo=FALSE, warning=FALSE}
#Calc normalization factors
dat.pDC.filter2.norm <- calcNormFactors(dat.pDC.filter2)

#voom normalize
temp <- voomWithQualityWeights(dat.pDC.filter2.norm, 
     design=model.matrix(~ IL5 + virus,
                         data=dat.pDC.filter2.norm$samples),
     plot=FALSE, save.plot = TRUE)

MV1 <- data.frame(
  x = temp$voom.xy$x, 
  y = temp$voom.xy$y,
  linex = temp$voom.line$x, 
  liney = temp$voom.line$y) %>%

  ggplot() +
  geom_point(aes(x=x, y=y), size=0.5) +
  geom_path(aes(x=linex, y=liney), color="red") +
  theme_classic() +
  labs(x="log2( count size + 0.5 )", y="Sqrt (stdev)",
       title="All genes")
```

```{r echo=FALSE}
#Calc normalization factors
dat.pDC.filter2.abund.norm <- calcNormFactors(dat.pDC.filter2.abund)

#Voom normalize
dat.pDC.filter2.abund.norm.voom <- voomWithQualityWeights(dat.pDC.filter2.abund.norm, 
     design=model.matrix(~ IL5 + virus,
            data=dat.pDC.filter2.abund.norm$samples),
     plot=FALSE, save.plot = TRUE)

#plot mean-var trend
MV2 <- data.frame(
  x = dat.pDC.filter2.abund.norm.voom$voom.xy$x, 
  y = dat.pDC.filter2.abund.norm.voom$voom.xy$y,
  linex = dat.pDC.filter2.abund.norm.voom$voom.line$x, 
  liney = dat.pDC.filter2.abund.norm.voom$voom.line$y) %>%

  ggplot() +
  geom_point(aes(x=x, y=y), size=0.5) +
  geom_path(aes(x=linex, y=liney), color="red") +
  theme_classic() +
  labs(x="log2( count size + 0.5 )", y="Sqrt (stdev)",
       title="Abundant genes")
```

```{r echo=FALSE, fig.height=4, fig.width=8.5}
plot_grid(MV1, MV2)
```

```{r echo=FALSE}
genes <- nrow(dat.pDC.filter2.norm$genes)
genes.abund <- nrow(dat.pDC.filter2.abund.norm$genes)
```

This removes `r genes-genes.abund` (~ `r round((genes-genes.abund)/genes*100, digits=0)`%) genes. This sufficiently filters these data as indicated by a decreasing variance trend for lowly abundant genes.

## PCA

We see that previously questionable samples (labeled with libID) are sufficiently normalized to not appear as PCA outliers.

```{r echo=FALSE, warning=FALSE, message=FALSE}
PCA.all <- as.data.frame(dat.pDC.filter2.abund.norm.voom$E) %>% 
  t() %>% 
  #Calc PCA
  prcomp()

#Extract axes labels with percent contribution of PC
PC1.label <- paste("PC1 (", summary(PCA.all)$importance[2,1]*100, "%)", sep="")
PC2.label <- paste("PC2 (", summary(PCA.all)$importance[2,2]*100, "%)", sep="")

# Extract PC values
PCA.dat <- as.data.frame(PCA.all$x) %>% 
    rownames_to_column("libID") %>%
    # Select PCs for plotting
    dplyr::select(libID, PC1:PC3) %>% 
    # Merge with metadata
    left_join(meta.filter2, by="libID")

PCA1 <- ggplot(PCA.dat, aes(PC1, PC2, color=virus:IL5)) +
      geom_point(size=2, stroke=1.5) +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nvoom normalized abundant logCPM") +
      coord_fixed(ratio=1) +
      scale_shape_manual(values=c(21,16)) +
      guides(color=guide_legend(title.position="top", 
                                title.hjust = 0.5))+
      scale_color_manual(values = group.cols)

PCA2 <- ggplot(PCA.dat, aes(PC1, PC2, color=donorID)) +
      geom_point(size=2, stroke=1.5) +
      geom_text_repel(data=filter(PCA.dat, 
                                  libID %in% to_remove$libID),
                      aes(label=libID), show.legend = FALSE) +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nvoom normalized abundant logCPM") +
      coord_fixed(ratio=1) +
      scale_color_manual(values=samp.cols)
```

```{r PCA.outlier3, echo=FALSE, message=FALSE}
PCA.all.sd <- PCA.dat %>% 
  group_by(virus, IL5) %>% 
  #Calculate PC mean std deviation
  summarise(.groups="keep",
    PC1.mean = mean(PC1),
    PC1.sd = sd(PC1),
    PC2.mean = mean(PC2),
    PC2.sd = sd(PC2)) %>% 
  #Calculate +/- 3 sd limits
  mutate(
    PC1.min = PC1.mean-(3*PC1.sd),
    PC1.max = PC1.mean+(3*PC1.sd),
    PC2.min = PC2.mean-(3*PC2.sd),
    PC2.max = PC2.mean+(3*PC2.sd))

PCA.all.dat <- PCA.dat %>%   
  full_join(PCA.all.sd) %>% 
  #ID potential outliers
  mutate(col.group = ifelse(PC1 > PC1.max | PC1 < PC1.min |
                            PC2 > PC2.max | PC2 < PC2.min, 
                            "potential outlier", "okay"))

PCA3 <- PCA.all.dat %>%       
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(color=col.group),size=2) + 
  geom_text_repel(data=filter(PCA.all.dat,
                            col.group == "potential outlier"),
            aes(label=libID), show.legend = FALSE) +
  #Beautify
  theme_classic() +
  theme(legend.title = element_blank(),
        legend.position = "bottom") +
  labs(x=PC1.label, y=PC2.label, 
       title="pDC experiment\nvoom normalized abundant logCPM") +
  coord_fixed(ratio=1) +
  scale_color_manual(values = c("#969696","#b10026"))
```

```{r echo=FALSE, fig.height=10, fig.width=8.5}
plot_grid(PCA1,PCA2,PCA3, align = "hv", ncol=2, axis="l")
```

## Summary

The following final quality cutoffs were applied to the data:

* Library median CV coverage $\leq$ `r round(max(dat.pDC.filter2.abund.norm.voom$targets$median_cv_coverage), digits=2)`
* Library mapped reads with duplicates $\geq$ `r round(min(dat.pDC.filter2.abund.norm.voom$targets$mapped_reads_w_dups), digits=2)`
* Library total counts $\geq$ `r min(dat.pDC.filter2.abund.norm.voom$targets$total_sequences)`
* Libraries from donor AC1 due to loss of media sample in previous filtering
* Within 3 sd of group mean on PCA (*e.g.* not outliers)
* Genes at least 1 CPM in at least 3 samples

```{r echo=FALSE, fig.height=4, fig.width=8.5}
# Plot libraries
meta %>% 
  #Create color group for removed samples
  mutate(col.group = ifelse(libID %in% dat.pDC.filter2$samples$libID,
                            "Retained","Removed")) %>%

ggplot(aes(median_cv_coverage, mapped_reads_w_dups)) +
  geom_point(aes(color=col.group, shape=experiment), size=2) +
  geom_hline(yintercept = 0.85, lty=2) +
  geom_vline(xintercept = 0.85, lty=2) +
  scale_y_continuous(breaks = sort(c(seq(0,1,length.out=5), 0.85)),
                    limits = c(0,1)) +
  scale_x_continuous(breaks = sort(c(seq(0,1, length.out=5), 0.85)),
                     limits = c(0,1)) +
  #Beautify
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x="Median CV coverage", y="Mapped reads with duplicates",
       color="Total sequences",
       title="Median CV coverage vs. mapped duplicate reads")
```

```{r echo=FALSE, fig.height=4, fig.width=8.5}
meta %>% 
  #Create color group for removed samples
  mutate(col.group = ifelse(libID %in% dat.pDC.filter2$samples$libID,
                            "Retained","Removed")) %>%

ggplot(aes(x=reorder_within(libID, by=total_sequences,
                            within=experiment), 
           y=total_sequences, fill=col.group))  +
  geom_col() +
  # Facet by cell type
  facet_grid(~experiment, scales="free_x", space="free") +
  # Add cutoff line
  geom_hline(yintercept = 1E5, lty=2) +
  # Beautify
  theme_classic() +
  theme(axis.text.x = element_text(size=rel(0.75),
                                   angle = 90, hjust = 1),
        legend.position = "bottom") +
  labs(x="Library", y="Total aligned counts\n(Log scale)", fill="",
       title="Total aligned counts") +
  scale_x_reordered() +
  scale_y_continuous(trans = 'log10',
    breaks = trans_breaks("log10", function(x) 10^x),
    labels = trans_format("log10", math_format(10^.x)))
```

```{r echo=FALSE}
filter.dup <- meta %>% 
  filter(mapped_reads_w_dups <= 0.85) %>% 
  mutate(filter = "mapped dup reads")
filter.cv <- meta %>% 
  filter(median_cv_coverage >= 0.85)%>% 
  mutate(filter = "CV coverage")
filter.seq <- meta %>% 
  filter(total_sequences <= 1E5)%>% 
  mutate(filter = "total seqs")
filter.AC1 <- meta %>% 
  filter(donorID == "AC1" & virus != "none")%>% 
  mutate(filter = "no media-only sample")
filter.AC2 <- meta %>% 
  filter(donorID == "AC2")%>% 
  mutate(filter = "PCA outlier")

bind_rows(filter.dup,filter.cv,filter.seq,
          filter.AC1,filter.AC2) %>% 
  group_by(libID) %>% 
  mutate(filters= paste(filter, collapse = ", \n")) %>% 
  select(experiment, libID, donorID, 
         asthma,IL5,virus, filters) %>%
  distinct() %>% 
  arrange(experiment, donorID, asthma,IL5,virus) %>%

  kable(align=c("c","c","c","c","l")) %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE)  
```

This results in the following samples for statistical analysis.

```{r echo=FALSE}
meta.filter2 %>% 
  group_by(experiment, asthma,IL5,virus) %>% 
  tally() %>% 
  arrange(experiment, asthma,IL5,virus) %>%

  kable(align="c") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
  collapse_rows(1:2, valign="top") 
```

# Save clean data

Normalized dat objects.

```{r message=FALSE}
#Save R object
#Rename all pDC object
dat.pDC.voom <- dat.pDC.filter2.abund.norm.voom

#P259_1
dat.pDC.voom_1 <- dat.pDC.voom

dat.pDC.voom_1$targets <- dat.pDC.voom_1$targets %>% 
  rownames_to_column() %>% 
  filter(experiment == "P259_1") %>% 
  droplevels() %>% 
  column_to_rownames()

dat.pDC.voom_1$E <- as.data.frame(dat.pDC.voom_1$E) %>% 
  rownames_to_column() %>% 
  select(rowname, dat.pDC.voom_1$targets$libID) %>% 
  column_to_rownames()

#P259_2
dat.pDC.voom_2 <- dat.pDC.voom

dat.pDC.voom_2$targets <- dat.pDC.voom_2$targets %>% 
  rownames_to_column() %>% 
  filter(experiment == "P259_2") %>% 
  droplevels() %>% 
  column_to_rownames()

dat.pDC.voom_2$E <- as.data.frame(dat.pDC.voom_2$E) %>% 
  rownames_to_column() %>% 
  select(rowname, dat.pDC.voom_2$targets$libID) %>% 
  column_to_rownames()

save(dat.pDC.voom, dat.pDC.voom_1, dat.pDC.voom_2,
     file="data_clean/P259_pDC_clean.RData")
```

# R session

```{r}
sessionInfo()
```

***

### 2020-09-09 2.P259.pDC_model.selection.Rmd
init public manuscript repo

---
title: 'P259: Model selection'
subtitle: "Dendritic cells (pDC)"
author: "Kim Dill-McFarland, kadm@uw.edu"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '4'
editor_options:
  chunk_output_type: console
---
# Background

The purpose of this workflow is to identify differentially expressed genes in pDC. Several models are compared in order to determine the best way to extract the effects of Anti-IL5 therapy and eosinophil (EOS) supernatant over the much larger human rhinovirus (RV) signal.

# Setup
Load packages

```{r message=FALSE, warning=FALSE}
# Data manipulation and figures
library(tidyverse)
    # Multi-panel figures for ggplot
    library(cowplot)
    # Add p-values to plots
    library(ggpubr)
#Venn diagrams
library(venn)
#Linear models
library(limma)
#Define ggplot colors
group.cols <- c("none:none"="#dadaeb",
               "none:AntiIL5"="#9e9ac8",
               "none:EOS.supp"="#54278f",
               "HRV:none"="#c7e9c0",
               "HRV:AntiIL5"="#74c476",
               "HRV:EOS.supp"="#006d2c",
               "flu:none"="#fdae6b",
               "flu:AntiIL5"="#e6550d")
samp.cols <- c("AC1"="#969696",
               "AC2"="#a6cee3",
               "AC3"="#1f78b4",
               "AC4"="#b2df8a",
               "AC5"="#33a02c",
               "AT1"="#fb9a99",
               "AT2"="#e31a1c",               
               "AT3"="#fdbf6f",
               "AT4"="#ff7f00",
               "donor1"="#cab2d6",
               "donor2"="#6a3d9a",
               "donor3"="#ffff99",
               "donor4"="#b15928")

# Print tty table to knit file
library(knitr)
library(kableExtra)
```

Set seed

```{r}
set.seed(4389)
```

Scripts and functions

```{r}
#To extract and format p-value from lmFit
source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/limma.extract.pval.R")
#opposite of %in%
`%notin%` <- Negate(`%in%`)
```

# Load data

Load normalized edgeR data objects. RNA-seq data was cleaned and normalized in [data cleaning](1.P259.pDC_data.cleaning.html). This included filtering for median CV coverage, mapped duplicate reads, total aligned counts, PCA outliers, and rare genes.

```{r}
load("data_clean/P259_pDC_clean.RData")
```

This includes in the following pDC samples.

```{r echo=FALSE}
dat.pDC.voom$targets %>% 
  group_by(experiment, asthma,IL5,virus) %>% 
  tally() %>% 
  arrange(experiment, asthma,IL5,virus) %>%

  kable(align="c") %>% 
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
  collapse_rows(1:2, valign="top") 
```

# Variables of interest

**Main effects**

* `donorID`: `r class(dat.pDC.voom$targets$donorID)` for each donor (N = `r length(unique(dat.pDC.voom$targets$donorID))`)
* `IL5`: `r class(dat.pDC.voom$targets$IL5)` of patient AntiIL5 therapy or sample EOS supernatant treatment (`r paste(levels(dat.pDC.voom$targets$IL5), collapse=", ")`)
* `virus`: `r class(dat.pDC.voom$targets$virus)` of viral conditions (`r paste(levels(dat.pDC.voom$targets$virus), collapse=", ")`)

**Fixed effects**

* `Sex`: `r paste(unique(dat.pDC.voom$targets$Sex), collapse=", ")`
* `Age`: age, years (`r min(dat.pDC.voom$targets$Age, na.rm=TRUE)` - `r max(dat.pDC.voom$targets$Age, na.rm=TRUE)`)
    - Neither available for 2 of 4 donors in P259.1

## Explore fixed effects and possible confounding variables

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=8.5}
plot1 <- dat.pDC.voom$targets %>% 
  select(experiment,donorID,IL5,Age) %>% 
  distinct() %>%

  ggplot(aes(x=IL5, y=Age)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(aes(color=donorID), height=0,width=0.2, size=3) +
  stat_compare_means(method="wilcox.test") +
  facet_wrap(~experiment, scales = "free_x") +
  theme_classic() +
  scale_color_manual(values=samp.cols) +
  labs(x="Treatment")

#Calculate chi-square for sex
sex.p.1 <- dat.pDC.voom$targets %>% 
  filter(experiment == "P259_1") %>% 
  droplevels()
sex.p.1 <- table(sex.p.1$IL5, sex.p.1$Sex) %>% 
  chisq.test()

sex.p.2 <- dat.pDC.voom$targets %>% 
  filter(experiment == "P259_2") %>% 
  droplevels()
sex.p.2 <- table(sex.p.2$IL5, sex.p.2$Sex) %>% 
  chisq.test()

plot2 <- dat.pDC.voom$targets %>% 
  group_by(experiment, IL5, Sex) %>% 
  tally() %>%

  ggplot(aes(x=IL5, y=n, fill=Sex)) +
  geom_bar(position="fill", stat="identity") +
  facet_wrap(~experiment, scales = "free_x") +
  theme_classic() +
  labs(y="Proportion", title=paste("Sex\nP259.1 Chi-squared = ", 
                                   round(sex.p.1$p.value,3),
                                   "\nP259.2 Chi-squared = ", 
                                   round(sex.p.2$p.value,3),
                                   sep="")) +
  scale_fill_manual(values=c("#f1a340","#998ec3"), na.value="grey")+
  labs(x="Treatment")

plot_grid(plot1, plot2) 
```

Age and sex appear to be different between treatment groups but are technically not significantly different, likely as the result of the small sample size. Due to this, the small sample size, and missing data in P259.1, sex and age will not be used as co-variates.

# PCA

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.height=5, fig.width=8.5}
# Calculate PCA
PCA <- as.data.frame(dat.pDC.voom$E) %>% 
  t() %>% 
  prcomp()

PC1.label <- paste("PC1 (", summary(PCA)$importance[2,1]*100, "%)", sep="")
PC2.label <-paste("PC2 (", summary(PCA)$importance[2,2]*100, "%)", sep="")

# Extract PC values
PCA.dat <- as.data.frame(PCA$x) %>% 
  rownames_to_column("libID") %>%
  # Select PCs
  dplyr::select(libID, PC1:PC3) %>% 
  # Merge with metadata
  left_join(dat.pDC.voom$targets, by="libID")

PCA1 <- ggplot(PCA.dat, aes(PC1, PC2, color=virus:IL5, shape=experiment)) +
      geom_point(size=2, stroke=1.5)  +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nvoom normalized abundant logCPM") +
      coord_fixed(ratio=1) +
      scale_shape_manual(values=c(21,16)) +
      guides(color=guide_legend(title.position="top", 
                                title.hjust = 0.5))+
      scale_color_manual(values = group.cols)

PCA2 <- ggplot(PCA.dat, aes(PC1, PC2, color=donorID, shape=experiment)) +
      geom_point(size=2, stroke=1.5) +
      #Beautify
      theme_classic() +
      labs(x=PC1.label, y=PC2.label, 
           title="pDC experiment\nvoom normalized abundant logCPM") +
      coord_fixed(ratio=1) +
      scale_shape_manual(values=c(21,16)) +
      scale_color_manual(values=samp.cols)

plot_grid(PCA1,PCA2, align = "hv", ncol=2)
```

# P259.1 model comparison
## Run models
### Interaction model

`gene expression ~ EOS.supp * virus + (1|donor)`

```{r}
# Define model
model_1.interact <- model.matrix(~ IL5*virus,
                          data=dat.pDC.voom_1$targets)
    colnames(model_1.interact) <- c("(Intercept)", "EOS.supp",
                            "HRV", "EOS.supp:HRV")
#Block by donor
consensus.corr <- duplicateCorrelation(
  dat.pDC.voom_1$E,
  model_1.interact,
  block=dat.pDC.voom_1$targets$donorID)$consensus.correlation

consensus.corr

# Fit model to transformed count data. Calculate eBayes
efitQW_1.interact <- eBayes(
                lmFit(dat.pDC.voom_1$E, 
                      model_1.interact,
                      block=dat.pDC.voom_1$targets$donorID,
                      correlation=consensus.corr))
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#Extract p-values from results
extract.pval(model=model_1.interact,
             voom.dat=dat.pDC.voom_1$E, 
             eFit=efitQW_1.interact, 
             name="pval_1.interact",
             summary=TRUE,
             contrasts=FALSE)
```

Summary

```{r echo=FALSE}
pval_1.interact.summ  %>%

  kable(align=c("l","c","c","c","c","c","c"),
        col.names = c("Variable", 
                      "<0.05", "<0.1",
                      "<0.2", "<0.3", "<0.4",
                      "<0.5")) %>% 
  kable_styling(latex_options="HOLD_position") %>% 
  add_header_above(c(" ", "Total genes at FDR" = 6))
```

### Contrasts model

`gene expression ~ EOS.supp:virus + (1|donor)`

```{r}
# Define model
dat.pDC.voom_1$targets <- dat.pDC.voom_1$targets %>% 
  mutate(contrast=paste(IL5, virus, sep="_")) %>% 
  mutate(contrast=factor(contrast, 
                         levels = c("none_none","none_HRV",
                                 "EOS.supp_none","EOS.supp_HRV")))

model_1.contrast<- model.matrix(~ 0 + contrast,
                          data=dat.pDC.voom_1$targets)
    colnames(model_1.contrast) <- c(
                                 "none_none","none_HRV",
                                 "EOS.supp_none","EOS.supp_HRV")
#Block by donor  
consensus.corr <- duplicateCorrelation(
  dat.pDC.voom_1$E,
  model_1.contrast,
  block=dat.pDC.voom_1$targets$donorID)$consensus.correlation

consensus.corr

#Fit model
fitQW_1.contrast <- lmFit(dat.pDC.voom_1$E, 
                      model_1.contrast,
                      block=dat.pDC.voom_1$targets$donorID,
                      correlation=consensus.corr)
#Get contrasts
contrast.matrix <- makeContrasts(
  none_HRV-none_none, 
  EOS.supp_HRV-EOS.supp_none,
  EOS.supp_none-none_none,
  EOS.supp_HRV-none_HRV,
  levels=model_1.contrast)

efitQW_1.contrast <- eBayes(contrasts.fit(fitQW_1.contrast,
                                  contrast.matrix))
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#Extract p-values from results
extract.pval(model=model_1.contrast,
             voom.dat=dat.pDC.voom_1$E, 
             eFit=efitQW_1.contrast, 
             name="pval_1.contrast",
             summary=TRUE,
             contrasts=TRUE,
             contrast.mat=contrast.matrix)
```

Summary

```{r echo=FALSE}
pval_1.contrast.summ  %>%

  kable(align=c("l","c","c","c","c","c","c"),
        col.names = c("Variable", 
                      "<0.05", "<0.1",
                      "<0.2", "<0.3", "<0.4",
                      "<0.5")) %>% 
  kable_styling(latex_options="HOLD_position") %>% 
  add_header_above(c(" ", "Total genes at FDR" = 6))
```

#### Contrast selected genes

Genes of interest are those that change with virus (*e.g.* significant for virus in untreated and/or treated samples) AND with EOS supernatant in media OR HRV.

```{r echo=FALSE, fig.height=6, fig.width=8.5}
#Define FDR cutoffs to assess
fdr.list <- c(0.05,0.1,0.2,0.3,0.4,0.5)

par(mfrow = c(2, 3))

for(fdr in fdr.list){
  #Blank list for results
  venn.list <- list()
  #Signif for virus
  temp <- pval_1.contrast %>% 
          filter(group %in% c("none_HRV - none_none",
                              "EOS.supp_HRV - EOS.supp_none") & 
                   adj.P.Val <= fdr) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["RV-infected\nvs media"]] <- temp

  #Signif in media
  temp2 <- pval_1.contrast %>% 
          filter(group %in% c("EOS.supp_none - none_none") & 
                   adj.P.Val <= fdr) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)

  venn.list[["- RV\nEOS supernatant\nvs none"]] <- temp2

  #Signif in HRV
  temp3 <- pval_1.contrast %>% 
          filter(group %in% c("EOS.supp_HRV - none_HRV") & 
                   adj.P.Val <= fdr) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)

  venn.list[["+ RV\nEOS supernatant\nvs none"]] <- temp3

  venn(ilab=FALSE, zcolor = "style",ilcs=1.5,sncs=1.5,
     x=venn.list, box=FALSE)

  title(sub=paste("FDR < ", fdr, sep=""), line = -1, cex.sub=1.5)
}
```

### Delta model

`RV - media gene expression ~ EOS.supp + (1|donor)`

Calculate HRV - media counts.

```{r echo=FALSE, message=FALSE}
dat.pDC.voom_1.delta.E <- as.data.frame(dat.pDC.voom_1$E) %>% 
  #Longer format, libID to columns
  rownames_to_column() %>% 
  pivot_longer(-rowname, names_to="libID") %>% 
  #Merge with metadata
  left_join(select(dat.pDC.voom_1$targets, libID, 
                   donorID, IL5, virus), by = "libID") %>% 
  select(-libID) %>% 
  #Separate media vs HRV samples
  pivot_wider(names_from = virus) %>% 
  #Calculate delta
  group_by(rowname, donorID, IL5) %>% 
  summarise(value = HRV-none) %>% 
  drop_na(value) %>% 
  ungroup() %>% 
  #Rename with donor.EOS
  mutate(new.name = paste(donorID, IL5, sep="_")) %>% 
  select(-donorID, -IL5) %>% 
  arrange(new.name, rowname) %>% 
  pivot_wider(names_from = new.name) %>% 
  #Matrix
  column_to_rownames()
```

```{r echo=FALSE}
#Subset metadata
dat.pDC.voom_1.delta.tgts <- dat.pDC.voom_1$targets %>% 
  select(-lib.size, -norm.factors, -libID, -sample.weights) %>% 
  filter(virus=="HRV") %>% 
  mutate(new.name = paste(donorID, IL5, sep="_")) %>% 
  filter(new.name %in% colnames(dat.pDC.voom_1.delta.E)) %>% 
  arrange(new.name)
```

Fit model

```{r}
# Define model
model_1.delta <- model.matrix(~ IL5,
                          data=dat.pDC.voom_1.delta.tgts)
    colnames(model_1.delta) <- c("(Intercept)","EOS.supp")

#Block by donor
consensus.corr <- duplicateCorrelation(
  dat.pDC.voom_1.delta.E,
  model_1.delta,
  block=dat.pDC.voom_1.delta.tgts$donorID)$consensus.correlation

consensus.corr

# Fit model to transformed count data. Calculate eBayes
efitQW_1.delta <- eBayes(
                lmFit(dat.pDC.voom_1.delta.E, 
                      model_1.delta,
                      block=dat.pDC.voom_1.delta.tgts$donorID,
                      correlation=consensus.corr))
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#Extract p-values from results
extract.pval(model=model_1.delta,
             voom.dat=dat.pDC.voom_1.delta.E, 
             eFit=efitQW_1.delta, 
             name="pval_1.delta",
             summary=TRUE,
             contrasts=FALSE)
```

Summary

```{r echo=FALSE}
pval_1.delta.summ %>%

  kable(align=c("l","c","c","c","c","c","c"),
        col.names = c("Variable", 
                      "<0.05", "<0.1",
                      "<0.2", "<0.3", "<0.4",
                      "<0.5")) %>% 
  kable_styling(latex_options="HOLD_position") %>% 
  add_header_above(c(" ", "Total genes at FDR" = 6))
```

## Compare models

```{r echo=FALSE}
fdr.cutoff <- 0.2
```

Total genes significant in one or more models at FDR < `r fdr.cutoff`

```{r echo=FALSE}
venn.list <- list()

#Signif for interaction
  temp <- pval_1.interact %>% 
          filter(group %in% c("EOS.supp:HRV") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["Signif interaction"]] <- temp

#Signif for delta
  temp <- pval_1.delta %>% 
          filter(group %in% c("EOS.supp") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["Signif delta"]] <- temp

#Signif for contrasts
  ##Signif for virus
  temp1 <- pval_1.contrast %>% 
          filter(group %in% c("none_HRV - none_none", 
                              "EOS.supp_HRV - EOS.supp_none") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  ##Signif for treatment
  temp2 <- pval_1.contrast %>% 
          filter(group %in% c("EOS.supp_none - none_none", 
                              "EOS.supp_HRV - none_HRV") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  temp <- pval_1.contrast %>% 
          filter(geneName %in% temp1 & geneName %in% temp2) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["Signif contrast"]] <- temp

par(mfrow = c(1,1)) 
  venn(ilab=FALSE, zcolor = "style",ilcs=1,sncs=1.5,
     x=venn.list)

  title(sub="FDR < 0.2", line = -1, cex.sub=1.5)
```

## Save results

```{r}
dir.create("results/gene_level/P259.1_model.selection/", 
           showWarnings = FALSE, recursive = TRUE)

save(pval_1.interact, pval_1.contrast, pval_1.delta,
     file="results/gene_level/P259.1_model.selection/P259.1_gene_pval.RData")

#gene lists
save(venn.list, file="results/gene_level/P259.1_model.selection/P259.1_gene_lists.RData")

#pval csv
pval_1.interact %>% 
  mutate(model = "interaction") %>% 
  bind_rows(mutate(pval_1.contrast, 
                   model="contrasts")) %>% 
  bind_rows(mutate(pval_1.delta, 
                   model="delta")) %>% 
  #Add HGNC symbol
  left_join(dat.pDC.voom_1$genes) %>% 
  write_csv(path = "results/gene_level/P259.1_gene_pval.csv")
```

# P259.2 model comparison
## Run models
### Interaction model

`gene expression ~ AntiIL5 * virus + (1|donor)`

```{r}
# Define model
model_2.interact <- model.matrix(~ IL5*virus,
                          data=dat.pDC.voom_2$targets)
    colnames(model_2.interact) <- c("(Intercept)", "AntiIL5",
                            "HRV", "AntiIL5:HRV")
#Block by donor
consensus.corr <- duplicateCorrelation(
  dat.pDC.voom_2$E,
  model_2.interact,
  block=dat.pDC.voom_2$targets$donorID)$consensus.correlation

consensus.corr

# Fit model to transformed count data. Calculate eBayes
efitQW_2.interact <- eBayes(
                lmFit(dat.pDC.voom_2$E, 
                      model_2.interact,
                      block=dat.pDC.voom_2$targets$donorID,
                      correlation=consensus.corr))
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#Extract p-values from results
extract.pval(model=model_2.interact,
             voom.dat=dat.pDC.voom_2$E, 
             eFit=efitQW_2.interact, 
             name="pval_2.interact",
             summary=TRUE,
             contrasts=FALSE)
```

Summary

```{r echo=FALSE}
pval_2.interact.summ  %>%

  kable(align=c("l","c","c","c","c","c","c"),
        col.names = c("Variable", 
                      "<0.05", "<0.1",
                      "<0.2", "<0.3", "<0.4",
                      "<0.5")) %>% 
  kable_styling(latex_options="HOLD_position") %>% 
  add_header_above(c(" ", "Total genes at FDR" = 6))
```

### Contrasts model

`gene expression ~ AntiIL5:virus + (1|donor)`

```{r}
# Define model
dat.pDC.voom_2$targets <- dat.pDC.voom_2$targets %>% 
  mutate(contrast=paste(IL5,virus, sep="_")) %>% 
  mutate(contrast=factor(contrast, 
                         levels = c("none_none","none_HRV",
                                 "AntiIL5_none","AntiIL5_HRV")))

model_2.contrast<- model.matrix(~ 0 + contrast,
                          data=dat.pDC.voom_2$targets)
    colnames(model_2.contrast) <- c(
                                 "none_none","none_HRV",
                                 "AntiIL5_none","AntiIL5_HRV")
#Block by donor  
consensus.corr <- duplicateCorrelation(
  dat.pDC.voom_2$E,
  model_2.contrast,
  block=dat.pDC.voom_2$targets$donorID)$consensus.correlation

consensus.corr

#Fit model
fitQW_2.contrast <- lmFit(dat.pDC.voom_2$E, 
                      model_2.contrast,
                      block=dat.pDC.voom_2$targets$donorID,
                      correlation=consensus.corr)

#Get contrasts
contrast.matrix <- makeContrasts(
  none_HRV-none_none, 
  AntiIL5_HRV-AntiIL5_none,
  AntiIL5_none-none_none,
  AntiIL5_HRV-none_HRV,
  levels=model_2.contrast)

efitQW_2.contrast <- eBayes(contrasts.fit(fitQW_2.contrast,
                                  contrast.matrix))
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#Extract p-values from results
extract.pval(model=model_2.contrast,
             voom.dat=dat.pDC.voom_2$E, 
             eFit=efitQW_2.contrast, 
             name="pval_2.contrast",
             summary=TRUE,
             contrasts=TRUE,
             contrast.mat=contrast.matrix)
```

Summary

```{r echo=FALSE}
pval_2.contrast.summ  %>%

  kable(align=c("l","c","c","c","c","c","c"),
        col.names = c("Variable", 
                      "<0.05", "<0.1",
                      "<0.2", "<0.3", "<0.4",
                      "<0.5")) %>% 
  kable_styling(latex_options="HOLD_position") %>% 
  add_header_above(c(" ", "Total genes at FDR" = 6))
```

### Contrast selected genes

Genes of interest are those that change with virus (*e.g.* significant for virus in untreated and/or treated donors) AND are different between untreated and treated donors in media OR RV.

```{r echo=FALSE, fig.height=6, fig.width=8.5}
#Define FDR cutoffs to assess
fdr.list<- c(0.05,0.1,0.2,0.3,0.4,0.5)

par(mfrow = c(2, 3))

for(fdr in fdr.list){
  #Blank list for results
  venn.list <- list()
  #Signif for virus
  temp <- pval_2.contrast %>% 
          filter(group %in% c("none_HRV - none_none",
                              "AntiIL5_HRV - AntiIL5_none") & 
                   adj.P.Val <= fdr) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["RV-infected\nvs media"]] <- temp

  #Signif in media
  temp2 <- pval_2.contrast %>% 
          filter(group %in% c("AntiIL5_none - none_none") & 
                   adj.P.Val <= fdr) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)

  venn.list[["- RV\nAntiIL5 therapy\nvs none"]] <- temp2

  #Signif in HRV
  temp3 <- pval_2.contrast %>% 
          filter(group %in% c("AntiIL5_HRV - none_HRV") & 
                   adj.P.Val <= fdr) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)

  venn.list[["+ RV\nAntiIL5 therapy\nvs none"]] <- temp3

  venn(ilab=FALSE, zcolor = "style",ilcs=1.5,sncs=1.5,
     x=venn.list, box=FALSE)

  title(sub=paste("FDR < ", fdr, sep=""), line = -1, cex.sub=1.5)
}
```

### Delta model

`RV - media gene expression ~ AntiIL5`

Calculate HRV - media counts.

```{r echo=FALSE, message=FALSE}
dat.pDC.voom_2.delta.E <- as.data.frame(dat.pDC.voom_2$E) %>% 
  #Longer format, libID to columns
  rownames_to_column() %>% 
  pivot_longer(-rowname, names_to="libID") %>% 
  #Merge with metadata
  left_join(select(dat.pDC.voom_2$targets, libID, 
                   donorID, virus.detail), by = "libID") %>% 
  select(-libID) %>% 
  #Separate media vs HRV samples
  pivot_wider(names_from = virus.detail) %>% 
  #Calculate delta
  group_by(rowname, donorID) %>% 
  summarise(oldHRV = oldHRV-none,
            newHRV = newHRV-none) %>% 
  #Rename with donor.HRV
  pivot_longer(-c(rowname, donorID)) %>% 
  mutate(new.name = paste(donorID, name, sep=".")) %>% 
  #Wide format
  select(rowname, new.name, value) %>% 
  arrange(new.name, rowname) %>% 
  pivot_wider(names_from = new.name) %>% 
  #Matrix
  column_to_rownames()
```

```{r echo=FALSE}
#Subset metadata
dat.pDC.voom_2.delta.tgts <- dat.pDC.voom_2$targets %>% 
  select(-lib.size, -norm.factors, -libID, -sample.weights) %>% 
  mutate(new.name = paste(donorID, virus.detail, sep=".")) %>% 
  filter(new.name %in% colnames(dat.pDC.voom_2.delta.E)) %>% 
  arrange(new.name)
```

Fit model

```{r}
# Define model
model_2.delta <- model.matrix(~ IL5,
                          data=dat.pDC.voom_2.delta.tgts)
    colnames(model_2.delta) <- c("(Intercept)","AntiIL5")

#Block by donor
consensus.corr <- duplicateCorrelation(
  dat.pDC.voom_2.delta.E,
  model_2.delta,
  block=dat.pDC.voom_2.delta.tgts$donorID)$consensus.correlation

consensus.corr

# Fit model to transformed count data. Calculate eBayes
efitQW_2.delta <- eBayes(
                lmFit(dat.pDC.voom_2.delta.E, 
                      model_2.delta,
                      block=dat.pDC.voom_2.delta.tgts$donorID,
                      correlation=consensus.corr))
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
#Extract p-values from results
extract.pval(model=model_2.delta,
             voom.dat=dat.pDC.voom_2.delta.E, 
             eFit=efitQW_2.delta, 
             name="pval_2.delta",
             summary=TRUE,
             contrasts=FALSE)
```

Summary

```{r echo=FALSE}
pval_2.delta.summ %>%

  kable(align=c("l","c","c","c","c","c","c"),
        col.names = c("Variable", 
                      "<0.05", "<0.1",
                      "<0.2", "<0.3", "<0.4",
                      "<0.5")) %>% 
  kable_styling(latex_options="HOLD_position") %>% 
  add_header_above(c(" ", "Total genes at FDR" = 6))
```

## Compare models

Total genes significant in one or more models at FDR < `r fdr.cutoff`

```{r echo=FALSE}
venn.list <- list()

#Signif for interaction
  temp <- pval_2.interact %>% 
          filter(group %in% c("AntiIL5:HRV") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["Signif interaction"]] <- temp

#Signif for delta
  temp <- pval_2.delta %>% 
          filter(group %in% c("AntiIL5") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["Signif delta"]] <- temp

#Signif for contrasts
  ##Signif for virus
  temp1 <- pval_2.contrast %>% 
          filter(group %in% c("none_HRV - none_none", 
                              "AntiIL5_HRV - AntiIL5_none") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  ##Signif for treatment
  temp2 <- pval_2.contrast %>% 
          filter(group %in% c("AntiIL5_none - none_none", 
                              "AntiIL5_HRV - none_HRV") & 
                   adj.P.Val <= fdr.cutoff) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  temp <- pval_2.contrast %>% 
          filter(geneName %in% temp1 & geneName %in% temp2) %>% 
          distinct(geneName) %>% 
          select(geneName) %>% unlist(use.names = FALSE)
  venn.list[["Signif contrast"]] <- temp

par(mfrow = c(1,1)) 
  venn(ilab=FALSE, zcolor = "style",ilcs=1,sncs=1.5,
     x=venn.list)

  title(sub="FDR < 0.5", line = -1, cex.sub=1.5)
```

## Save results

```{r}
dir.create("results/gene_level/P259.2_model.selection/", 
           showWarnings = FALSE, recursive = TRUE)

save(pval_2.interact, pval_2.contrast, pval_2.delta,
     file="results/gene_level/P259.2_model.selection/P259.2_gene_pval.RData")

#gene lists
save(venn.list, file="results/gene_level/P259.2_model.selection/P259.2_gene_lists.RData")

#pval csv
pval_2.interact %>% 
  mutate(model = "interaction") %>% 
  bind_rows(mutate(pval_2.contrast, 
                   model="contrasts")) %>% 
  bind_rows(mutate(pval_2.delta, 
                   model="delta")) %>% 
  #Add HGNC symbol
  left_join(dat.pDC.voom_2$genes) %>% 
  write_csv(path = "results/gene_level/P259.2_gene_pval.csv")
```

# R session

```{r}
sessionInfo()
```

***

### 2020-09-09 3.P259.pDC_GSEA.Rmd
init public manuscript repo

---
title: 'P259: Gene set enrichment analysis (GSEA)'
subtitle: "Dendritic cells (pDC)"
author: "Kim Dill-McFarland, kadm@uw.edu"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
  pdf_document:
    toc: yes
    toc_depth: '3'
editor_options:
  chunk_output_type: console
---
# Background

The purpose of this workflow is to perform GSEA for the impacts of human rhinovirus (RV) infection, eosinophil (EOS) supernatant, and Anti-IL5 therapy.

# Setup
Load packages

```{r message=FALSE, warning=FALSE}
# Data manipulation and figures
library(tidyverse)
# Multipanel figures
library(cowplot)

#GSEA
library(fgsea)
library(gage)

#Print pretty tables to Rmd
library(knitr)
library(kableExtra)
```

Set seed

```{r}
set.seed(589)
```

# Load data
## RNA-seq

Contrast model results.

```{r results.data, message=FALSE}
pval_2 <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
  filter(model=="contrasts")
pval_1 <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
  filter(model=="contrasts")
```

Extract and format fold change (FC) for each contrast.

```{r}
gene.ls <- list()

for (contrast in unique(pval_2$group)){
  #Subset to contrast of interest
  pval.temp <- pval_2 %>% filter(group == contrast)

  genes.temp <- pval.temp$logFC
  names(genes.temp) <- pval.temp$hgnc_symbol

  list.name <- paste(gsub(" - ", ".", contrast), 2, sep=".")
  gene.ls[[list.name]] <- genes.temp
}

for (contrast in unique(pval_1$group)){
  #Subset to contrast of interest
  pval.temp <- pval_1 %>% filter(group == contrast)

  genes.temp <- pval.temp$logFC
  names(genes.temp) <- pval.temp$hgnc_symbol

  list.name <- paste(gsub(" - ", ".", contrast), 1, sep=".")
  gene.ls[[list.name]] <- genes.temp
}
```

## Broad gene sets

From <https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp>. Downloaded in `data_clean/Broad_gmt/`

# Gene set enrichment analysis (GSEA)

The following function performs GSEA using both fast gene set enrichment analysis (`fgsea`) and generally applicable gene-set enrichment (`gage`).

```{r}
source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/GSEA_fxn.R")
```

## Run GSEA
#### Hallmark (H)

```{r h, warnings=FALSE, message=FALSE, eval=FALSE}
GSEA(gene_list = gene.ls, 
     gmt_file="data_clean/Broad_gmt/h.all.v7.1.symbols.gmt")
```

#### Curated gene sets (C2)

```{r c2, warnings=FALSE, message=FALSE, eval=FALSE}
GSEA(gene_list = gene.ls, 
     gmt_file="data_clean/Broad_gmt/c2.all.v7.1.symbols.gmt")
```

#### GO gene sets (C5)

```{r c5, warnings=FALSE, message=FALSE, eval=FALSE}
GSEA(gene_list = gene.ls, 
     gmt_file="data_clean/Broad_gmt/c5.all.v7.1.symbols.gmt")
```

## Significant GSEA

```{r echo=FALSE, message=FALSE}
#Load results for faster knitting
h_GSEA.result <- read_csv("results/GSEA_FoldChange/h_GSEA.result.csv")
c2_GSEA.result <- read_csv("results/GSEA_FoldChange/c2_GSEA.result.csv")
c5_GSEA.result <- read_csv("results/GSEA_FoldChange/c5_GSEA.result.csv")
```

Gene sets of interest are those significant for both virus AND EOS supernatant or Anti-IL5 therapy. Results are only considered significant if both fgsea and gage methods meet the FDR threshold in the same fold change direction.

```{r echo=FALSE}
source("scripts/GSEA.score.plot.R")
```

#### Hallmark FDR < 0.1

```{r echo=FALSE, fig.height=5, fig.width=8.5}
gsea.plot(gsea = h_GSEA.result, fdr.cut = 0.1, 
          prefix = "H", print.plot = TRUE, save.plot = FALSE)
```

#### Curated gene sets (C2) FDR < 0.05

```{r echo=FALSE, fig.height=25, fig.width=8.5}
gsea.plot(gsea = c2_GSEA.result, fdr.cut = 0.05, 
          prefix = "C2", print.plot = TRUE, save.plot = FALSE)
```

#### GO gene sets (C5) FDR < 0.05

```{r echo=FALSE, fig.height=25, fig.width=8.5}
gsea.plot(gsea = c5_GSEA.result, fdr.cut = 0.05, 
          prefix = "C5", print.plot = TRUE, save.plot = FALSE)
```

# R session

```{r}
sessionInfo()
```

***

### 2020-09-09 scripts/GSEA.score.plot.R

@@ -1,0 +1,129 @@
+gsea.plot <- function(gsea, fdr.cut = 0.1, prefix,
+                      print.plot = TRUE, save.plot = FALSE, 
+                      height = 5, width = 11, outdir=NULL){
+  require(tidyverse)
+  require(cowplot)
+  
+  #### Format data ####
+  #list all terms with at least 1 significant treatment contrast
+  term.OI.t <- gsea %>% 
+    filter(fgsea.FDR <= fdr.cut & gage.FDR <= fdr.cut) %>% 
+    filter(group %in% c("AntiIL5_none.none_none.2", "AntiIL5_HRV.none_HRV.2",
+                        "EOS.supp_none.none_none.1", "EOS.supp_HRV.none_HRV.1")) %>% 
+    distinct(pathway) %>% unlist(use.names=FALSE)
+  
+  #list all terms with at least 1 significant virus contrast 
+  term.OI.v <- gsea %>% 
+    filter(fgsea.FDR <= fdr.cut & gage.FDR <= fdr.cut) %>% 
+    filter(group %in% c("none_HRV.none_none.2","AntiIL5_HRV.AntiIL5_none.2",
+                        "none_HRV.none_none.1","EOS.supp_HRV.EOS.supp_none.1")) %>% 
+    distinct(pathway) %>% unlist(use.names=FALSE)
+  
+  #Find terms with both t and v cutoffs
+  term.OI <- intersect(term.OI.t, term.OI.v)
+  
+  h.plot.dat <- gsea %>% 
+    filter(pathway %in% term.OI) %>% 
+    mutate(group1 = recode_factor(factor(group),
+                                  "AntiIL5_HRV.AntiIL5_none.2"="+ Anti-IL5 therapy",
+                                  "none_HRV.none_none.2"="- Anti-IL5 therapy",
+                                  "EOS.supp_HRV.EOS.supp_none.1"="+ EOS supernatant",
+                                  "none_HRV.none_none.1"="- EOS supernatant",
+                                  
+                                  "EOS.supp_HRV.none_HRV.1"="+ RV",
+                                  "AntiIL5_HRV.none_HRV.2"="+ RV",
+                                  "EOS.supp_none.none_none.1"="- RV",
+                                  "AntiIL5_none.none_none.2"="- RV"),
+           group2 = recode_factor(factor(group),
+                                  "none_HRV.none_none.1"="RV-infected vs media",
[…]

### 2021-05-10 publication/4.P259.pDC_publication.fig.table.Rmd
Mv pub rmd to publication dir

---
title: 'P259: Publication figures and tables'
subtitle: "Dendritic cells (pDC)"
author: "Kim Dill-McFarland, kadm@uw.edu"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
---

```{r message=FALSE, warning=FALSE}
#Create directories of outputs
dir.create("publication/fig/", showWarnings = FALSE, recursive = TRUE)
dir.create("publication/table/", showWarnings = FALSE, recursive = TRUE)

#Load packages used across all data
library(tidyverse)
library(limma)
#Set seed
set.seed(4389)
```

# Tables
## RNA-seq library and patient metadata

```{r}
attach("data_clean/P259_pDC_clean.RData")

dat.pDC.voom$targets %>% 
  select(libID, everything()) %>% 
  arrange(libID) %>% 
  write_csv(., "publication/table/TableS.metadata.csv")
```

### Patient summary demographics

```{r}
#Age
dat.pDC.voom$targets %>% 
  distinct(experiment, donorID, Age) %>% 
  group_by(experiment) %>% 
  summarise(.groups = "keep",
            n = n(),
            mean.age = mean(Age, na.rm=TRUE),
            sd.age = sd(Age, na.rm=TRUE)/sqrt(n))
#Sex
dat.pDC.voom$targets %>% 
  distinct(experiment, donorID, Sex) %>% 
  count(experiment, Sex) %>% 
  group_by(experiment) %>% 
  mutate(pct = n/sum(n))
```

## RNA-seq normalized log2 counts

```{r}
#attach("data_clean/P259_pDC_clean.RData")

as.data.frame(dat.pDC.voom$E) %>% 
  rownames_to_column("geneName") %>% 
#Save
write_csv(., "publication/table/TableS.norm.log2.counts.csv")
```

## All contrast model results

```{r message=FALSE}
#Get all model results
read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
  #Add experiment variable
  mutate(experiment = "P259_1") %>% 
  bind_rows(read_csv("results/gene_level/P259.2_gene_pval.csv")) %>% 
  #Fill in experiment variable
  mutate(experiment = ifelse(is.na(experiment), "P259_2", experiment)) %>% 
  #Keep only contrasts model 
  filter(model == "contrasts") %>% 
  #Reorder and rename variables
  select(experiment, model, geneName, hgnc_symbol, group,
         AveExpr:group) %>% 
  rename(contrast = group) %>% 
  arrange(experiment, geneName, contrast) %>% 
#Save
write_csv(., "publication/table/TableS.models.csv")
```

## GSEA results

```{r message=FALSE}
read_csv("results/GSEA_FoldChange/h_GSEA.result.csv") %>% 
  rename(contrast = group) %>% 
  #Add experiment variable
  mutate(experiment = ifelse(grepl(".1", contrast), "P259_1", "P259_2")) %>%
  #Set gene set variable
  mutate(gene.set = "HALLMARK",
         pathway = gsub("HALLMARK_", "", pathway)) %>% 
  #Format contrast to match model results
  mutate(contrast = gsub(".1", "", contrast),
         contrast = gsub(".2", "", contrast),
         contrast = gsub("[.]n"," - n", contrast),
         contrast = gsub("[.]A"," - A", contrast),
         contrast = gsub("[.]E"," - E", contrast)) %>% 
  #Reorder variables
  select(experiment, gene.set, pathway, contrast, everything()) %>% 
  arrange(experiment, pathway, contrast) %>% 
#Save
write_csv(., "publication/table/TableS.GSEA.csv") 
```

# Figures
## PCA

```{r message=FALSE}
library(cowplot)
```

Calculate PCA for P259.1 and P259.2

```{r}
#attach("data_clean/P259_pDC_clean.RData")

PCA1 <- as.data.frame(dat.pDC.voom_1$E) %>% 
  t() %>% 
  prcomp(scale. = TRUE)

#Extract axes labels
PC1.label1 <- paste("PC1 (", 
                    round(summary(PCA1)$importance[2,1]*100, digits=1), 
                    "%)", sep="")
PC2.label1 <-paste("PC2 (", 
                   round(summary(PCA1)$importance[2,2]*100, digits=1), 
                   "%)", sep="")

PCA2 <- as.data.frame(dat.pDC.voom_2$E) %>% 
  t() %>% 
  prcomp(scale. = TRUE)

#Extract axes labels
PC1.label2 <- paste("PC1 (", 
                    round(summary(PCA2)$importance[2,1]*100, digits=1), 
                    "%)", sep="")
PC2.label2 <-paste("PC2 (", 
                   round(summary(PCA2)$importance[2,2]*100, digits=1), 
                   "%)", sep="")
```

Combine experiments and metadata.

```{r}
# Extract PC values
PCA.dat <- as.data.frame(PCA1$x) %>% 
  rownames_to_column("libID") %>%
  bind_rows(rownames_to_column(as.data.frame(PCA2$x),"libID")) %>% 
  # Select PCs
  dplyr::select(libID, PC1, PC2) %>% 
  # Merge with metadata
  left_join(dat.pDC.voom$targets, by="libID")
```

Plot PCAs

```{r fig.width=8.5}
#Define ggplot colors
group.cols <- c("none:none"="#dadaeb",
                "none:AntiIL5"="#9e9ac8",
                "none:EOS.supp"="#54278f",
                "HRV:none"="#c7e9c0",
                "HRV:AntiIL5"="#74c476",
                "HRV:EOS.supp"="#006d2c",
                "flu:none"="#fdae6b",
                "flu:AntiIL5"="#e6550d")

PCA1 <- PCA.dat %>% 
  filter(experiment=="P259_1") %>%

  ggplot(aes(PC1, PC2, color=virus:IL5)) +
  geom_point(size=5) +
  #Beautify
  theme_classic(base_size = 16) +
  labs(x=PC1.label1, y=PC2.label1) +
  coord_fixed(ratio=1)  +
  scale_color_manual(values = group.cols, 
                     labels = c("media", "+ EOS sup", 
                              "+ RV", "+ RV + EOS sup"),
                     name="") +
  theme(legend.position = "bottom") +
  guides(color=guide_legend(nrow=2))

PCA2 <- PCA.dat %>% 
  filter(experiment=="P259_2") %>%

  ggplot(aes(PC1, PC2, color=virus:IL5)) +
  geom_point(size=5) +
  #Beautify
  theme_classic(base_size = 16) +
  labs(x=PC1.label2, y=PC2.label2) +
  coord_fixed(ratio=1) +
  scale_color_manual(values = group.cols, 
                     labels = c("media", "+ Anti-IL5", 
                                "+ RV", "+ RV + Anti-IL5"),
                     name="") +
  theme(legend.position = "bottom") +
  guides(color=guide_legend(nrow=2))

plot_grid(PCA1,PCA2, align = "hv", labels = c("A","B"))

#Save
ggsave("publication/fig/PCA.pdf", 
       plot_grid(PCA1,PCA2, align = "hv", labels = c("A","B")),
       width=10, height=5)
```

## Contrast model venn

```{r}
library(venn)
```

Genes significant for contrasts at FDR < 0.2

```{r message=FALSE, fig.width=8.5}
#Load pvalue results
pval_1 <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
  filter(model == "contrasts")
pval_2 <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
  filter(model == "contrasts")

#Set FDR cutoff
fdr.list <- c(0.2)

#Plot venns
for(fdr.cutoff in fdr.list){
  #P259.1
  venn.list_1 <- list()
  ##Signif for virus
  venn.list_1[["RV-infected\nvs media"]] <- pval_1 %>% 
    filter(group %in% c("none_HRV - none_none",
                        "EOS.supp_HRV - EOS.supp_none") & 
             adj.P.Val <= fdr.cutoff) %>% 
    distinct(geneName) %>% 
    select(geneName) %>% unlist(use.names = FALSE)

  ##Signif for treatment
  venn.list_1[["- RV\nEOS supernatant\nvs none"]] <- pval_1 %>% 
    filter(group == "EOS.supp_none - none_none" & 
             adj.P.Val <= fdr.cutoff) %>% 
    distinct(geneName) %>% 
    select(geneName) %>% unlist(use.names = FALSE)

  venn.list_1[["+ RV\nEOS supernatant\nvs none"]] <- pval_1 %>% 
    filter(group == "EOS.supp_HRV - none_HRV" & 
             adj.P.Val <= fdr.cutoff) %>% 
    distinct(geneName) %>% 
    select(geneName) %>% unlist(use.names = FALSE)

  #P259.2
  venn.list_2 <- list()
  ##Signif for virus
  venn.list_2[["RV-infected\nvs media"]] <- pval_2 %>% 
    filter(group %in% c("none_HRV - none_none",
                        "AntiIL5_HRV - AntiIL5_none") & 
             adj.P.Val <= fdr.cutoff) %>% 
    distinct(geneName) %>% 
    select(geneName) %>% unlist(use.names = FALSE)

  ##Signif for treatment
  venn.list_2[["- RV\nAnti-IL5 therapy\nvs none"]] <- pval_2 %>% 
    filter(group == "AntiIL5_none - none_none" & 
             adj.P.Val <= fdr.cutoff) %>% 
    distinct(geneName) %>% 
    select(geneName) %>% unlist(use.names = FALSE)

  venn.list_2[["+ RV\nAnti-IL5 therapy\nvs none"]] <- pval_2 %>% 
    filter(group == "AntiIL5_HRV - none_HRV" & 
             adj.P.Val <= fdr.cutoff) %>% 
    distinct(geneName) %>% 
    select(geneName) %>% unlist(use.names = FALSE)

  #Save
  pdf(paste("publication/fig/contrast.venn_", fdr.cutoff, ".pdf", sep=""),
      height=5, width=12)
  par(mfrow = c(1,2)) 
  venn(ilab=FALSE, zcolor = "style",ilcs=1.5, sncs=1.5, box=FALSE,
       x=venn.list_1)
  venn(ilab=FALSE, zcolor = "style",ilcs=1.5, sncs=1.5, box=FALSE,
       x=venn.list_2)
  dev.off()

  #Print to Rmd
  par(mfrow = c(1,2)) 
  venn(ilab=FALSE, zcolor = "style",ilcs=1.5, sncs=1.5, box=FALSE,
       x=venn.list_1)
  venn(ilab=FALSE, zcolor = "style",ilcs=1.5, sncs=1.5, box=FALSE,
       x=venn.list_2)
}
```

## Hallmark GSEA 
### Enrichment score

```{r message=FALSE, fig.width=8.5}
#Load plotting function
source("scripts/GSEA.score.plot.R")

#Create plot
gsea.plot(gsea = read_csv("results/GSEA_FoldChange/h_GSEA.result.csv"),
          fdr.cut = 0.1, prefix = "H", print.plot = TRUE, 
          save.plot = TRUE, height = 5, width = 11, 
          outdir="publication/fig/")
```

### Mean fold change

```{r}
library(cowplot)
library(ggrepel)
```

```{r message=FALSE, warning=FALSE, fig.width=8.5, fig.height=10}
#List genes in each Hallmark term
myGO <- fgsea::gmtPathways("data_clean/Broad_gmt/h.all.v7.1.symbols.gmt")
GO.df <- plyr::ldply(myGO, rbind) %>% 
  rename(term = `.id`) %>% 
  pivot_longer(-term, values_to="hgnc_symbol")

#Expression data
#attach("data_clean/P259_pDC_clean.RData")

#### Format data ####
#Select genes of interest and format
## P259.2
h.combo.plot_2 <- as.data.frame(dat.pDC.voom_2$E) %>% 
  rownames_to_column("geneName") %>% 
  #Get HGNC symbols
  left_join(dat.pDC.voom_2$genes) %>% 
  select(hgnc_symbol, everything()) %>% 
  #Filter to genes in terms
  inner_join(GO.df) %>% 
  select(term, hgnc_symbol, geneName, everything(), 
         -c(`Previous symbols`:gene_biotype), -name) %>% 
  #long format
  pivot_longer(-c(term:geneName), names_to = "libID", 
               values_to = "expression") %>% 
  #Add metadata
  mutate(experiment = "P259.2") %>% 
  left_join(select(dat.pDC.voom_2$targets, donorID, libID, 
                   IL5, virus.detail))

## P259.1
h.combo.plot_1 <- as.data.frame(dat.pDC.voom_1$E) %>% 
  rownames_to_column("geneName") %>% 
  #Get HGNC symbols
  left_join(dat.pDC.voom_1$genes) %>% 
  select(hgnc_symbol, everything()) %>% 
  #Filter to genes in terms
  inner_join(GO.df) %>% 
  select(term, hgnc_symbol, geneName, everything(), 
         -c(`Previous symbols`:gene_biotype), -name) %>% 
  #long format
  pivot_longer(-c(term:geneName), names_to = "libID", 
               values_to = "expression") %>% 
  #Add metadata
  mutate(experiment = "P259.1") %>% 
  left_join(select(dat.pDC.voom_1$targets, donorID,
                   libID, IL5, virus.detail))

#Combine data and calculate fold change
h.combo.plot <- bind_rows(h.combo.plot_1, h.combo.plot_2) %>% 
  mutate(group = paste(virus.detail, IL5, sep=".")) %>% 
  select(-IL5, -virus.detail, -libID) %>% 
  #Averages
  group_by(term, hgnc_symbol, experiment, group) %>% 
  summarise(meanE = mean(expression, na.rm=TRUE)) %>% 
  #Fold change
  pivot_wider(names_from = group, values_from = meanE) %>% 
  rowwise() %>% 
  #Simplify group names
  mutate(
    #Virus in untreated
    virus.untreat_old = oldHRV.none-none.none,
    virus.untreat_new = newHRV.none-none.none,
    #Virus in treated
    virus.treat_old1 = oldHRV.EOS.supp-none.EOS.supp,
    virus.treat_old2 = oldHRV.AntiIL5-none.AntiIL5,
    virus.treat_new = newHRV.AntiIL5-none.AntiIL5,
    #Treatment in media
    treat.media_1 = none.EOS.supp-none.none,
    treat.media_2 = none.AntiIL5-none.none,
    #Treatment in virus
    treat.virus_old1 = oldHRV.EOS.supp-oldHRV.none,
    treat.virus_old2 = oldHRV.AntiIL5-oldHRV.none,
    treat.virus_new = newHRV.AntiIL5-newHRV.none) %>% 
  ungroup() %>%

  #long format
  select(term:experiment, virus.untreat_old:treat.virus_new) %>% 
  pivot_longer(virus.untreat_old:treat.virus_new) %>% 
  drop_na(value) %>%

  #Average old and new HRV values
  separate(name, into=c("group"), sep="_") %>% 
  group_by(term, hgnc_symbol, experiment, group) %>% 
  summarise(value = mean(value, na.rm=TRUE)) %>% 
  ungroup() %>% 
  #Beautify for plotting
  mutate(term = gsub("HALLMARK_","", term),
         term = gsub("_"," ", term)) %>% 
  mutate(col.group = ifelse(value <0, "down", "up")) %>% 
  mutate(group = factor(group, levels=c("treat.media","treat.virus",
                                        "virus.untreat","virus.treat")))

#Set jitter
pos <- position_jitter(width = 0.3, seed = 589, height = 0)

#### List genes to label ####
#Set max number of terms
no.genes <- 5
#Set gene sets of interest
term.OI.ls <- c("INTERFERON ALPHA RESPONSE", "INTERFERON GAMMA RESPONSE",
             "INFLAMMATORY RESPONSE",
             "EPITHELIAL MESENCHYMAL TRANSITION")

#Find genes with largest absolute fold change  
to.label <- list()

for(term.OI in term.OI.ls){
  #If this term, find genes that go down with virus
  if(term.OI == "EPITHELIAL MESENCHYMAL TRANSITION"){
    genes.v <- h.combo.plot %>% 
      filter(term ==term.OI &
               group %in% c("virus.treat","virus.untreat")) %>% 
      group_by(term, hgnc_symbol) %>% 
      filter(max(value, na.rm=TRUE)<0) %>% 
      summarise(max.fc = max(abs(value), na.rm=TRUE)) %>% 
      ungroup() %>% 
      slice_max(max.fc,n=no.genes) %>% 
      distinct(hgnc_symbol)
  } else{
    #All other terms, find genes that go up with virus
    genes.v <- h.combo.plot %>% 
      filter(term ==term.OI &
               group %in% c("virus.treat","virus.untreat")) %>% 
      group_by(term, hgnc_symbol) %>% 
      filter(min(value, na.rm=TRUE)>0) %>% 
      summarise(max.fc = max(value, na.rm=TRUE)) %>% 
      ungroup() %>% 
      slice_max(max.fc,n=no.genes) %>% 
      distinct(hgnc_symbol)
  }

  #Save genes to list object
  to.label[[paste("v",term.OI,sep=".")]] <- genes.v

  #Find genes that go down with EOS supp
  down.eos <- h.combo.plot %>% 
    filter(term == term.OI & group %in% c("treat.media","treat.virus") &
             experiment == "P259.1") %>% 
    group_by(term, hgnc_symbol) %>% 
    filter(max(value, na.rm=TRUE)<0) %>% 
    ungroup() %>% 
    distinct(hgnc_symbol) %>% unlist(use.names = FALSE)

  #Find genes that go up with AntiIL5
  up.aIL5 <- h.combo.plot %>% 
    filter(term == term.OI & group %in% c("treat.media","treat.virus") &
             experiment == "P259.2") %>% 
    group_by(term, hgnc_symbol) %>% 
    filter(min(value, na.rm=TRUE)>0) %>% 
    ungroup() %>% 
    distinct(hgnc_symbol) %>% unlist(use.names = FALSE)

  #Concatenate to genes that go up/down with virus and down with EOS and 
  # up with AntiIL5
  ## i.e. follow the directions in the enrichment score plot
  genes.t <- h.combo.plot %>% 
    filter(term == term.OI &
             group %in% c("virus.treat","virus.untreat")) %>% 
    filter(hgnc_symbol %in% intersect(down.eos,up.aIL5)) %>%

    group_by(term, hgnc_symbol) %>% 
    filter(max(abs(value), na.rm=TRUE)>0) %>% 
    summarise(max.fc = max(abs(value), na.rm=TRUE)) %>% 
    ungroup() %>% 
    slice_max(max.fc,n=no.genes) %>% 
    distinct(hgnc_symbol)

  #Save to list object
  to.label[[paste("t",term.OI,sep=".")]] <- genes.t
}

#Convert genes to label to df
to.label.df <- data.table::rbindlist(to.label, idcol="name") %>% 
  separate(name, into=c("vt","term"), sep="[.]") %>% 
  mutate(group1 = ifelse(vt == "t", "treat.media", "virus.untreat"),
         group2 = ifelse(vt == "t", "treat.virus", "virus.treat")) %>% 
  pivot_longer(group1:group2, values_to="group") %>% 
  dplyr::select(-name, -vt) %>% 
  mutate(to.label = "y")

#add genes to label info to other  plot data
h.combo.plot.lab <- h.combo.plot %>% 
  full_join(to.label.df, by = c("term", "hgnc_symbol", "group"))

### FC plot data ####
FC.plot.dat <- h.combo.plot.lab %>% 
  #Beautify labels
  mutate(group = paste(experiment, group, sep="_")) %>% 
  mutate(group1 = recode_factor(factor(group),
                  "P259.2_virus.treat"="+ Anti-IL5 therapy",
                  "P259.2_virus.untreat"="- Anti-IL5 therapy",
                  "P259.1_virus.treat"="+ EOS supernatant",
                  "P259.1_virus.untreat"="- EOS supernatant",

                  "P259.1_treat.virus"="+ RV",
                  "P259.2_treat.virus"="+ RV",
                  "P259.1_treat.media"="- RV",
                  "P259.2_treat.media"="- RV"),
         group2 = recode_factor(factor(group),
                  "P259.1_virus.untreat"="RV-infected vs media",
                  "P259.2_virus.untreat"="RV-infected vs media",
                  "P259.1_virus.treat"="RV-infected vs media",
                  "P259.2_virus.treat"="RV-infected vs media",

                  "P259.1_treat.virus"="EOS supernatant vs none",
                  "P259.2_treat.virus"="Anti-IL5 therapy vs none",
                  "P259.1_treat.media"="EOS supernatant vs none",
                  "P259.2_treat.media"="Anti-IL5 therapy vs none")) %>% 
    #Reorder terms
    mutate(term.ord = recode_factor(factor(term), 
  "INTERFERON ALPHA RESPONSE"="INTERFERON\nALPHA\nRESPONSE",
  "INTERFERON GAMMA RESPONSE"="INTERFERON\nGAMMA\nRESPONSE",
  "INFLAMMATORY RESPONSE"="INFLAMMATORY\nRESPONSE",
  "EPITHELIAL MESENCHYMAL TRANSITION"="EPITHELIAL\nMESENCHYMAL\nTRANSITION"))

#### FC plot: VIRUS ####
dat.v <- FC.plot.dat %>% 
  #Reorder terms
      filter(group2 == "RV-infected vs media" &
             term.ord %in% c("INTERFERON\nALPHA\nRESPONSE",
                           "INTERFERON\nGAMMA\nRESPONSE",
                           "INFLAMMATORY\nRESPONSE",
                           "EPITHELIAL\nMESENCHYMAL\nTRANSITION"))

plot.v <- dat.v %>% 
      ggplot(aes(x = group1, y=value)) +
      geom_violin() +
      #Add non-labeled points
      geom_jitter(data=filter(dat.v, is.na(to.label)),
                  aes(color=col.group), position = pos) +
      #Add labeled points
      geom_point(data=filter(dat.v, !is.na(to.label)),
                 aes(color=col.group), ) +
      #Add labels left
      geom_text_repel(data=filter(dat.v, !is.na(to.label)),
                      aes(label=hgnc_symbol), direction="y",
                      hjust=1, nudge_x=-0.4,
                      show.legend = FALSE, size=3) +
      #Add line at 0
      #geom_hline(yintercept = 0, size=0.5) +
      #Add mean lines
      stat_summary(fun="mean", geom="crossbar")+
      facet_grid(term.ord~group2, scales="free") +
      coord_flip() +
      #Beautify
      theme_bw() +
      labs(x="", y="Gene mean log2 fold change") +
      scale_color_manual(values=c("down"="#6baed6","up"="#ef3b2c")) +
      theme(strip.text.y = element_blank())+
      theme(legend.position = "none",
            panel.grid.major.y = element_blank(),
            panel.grid.minor.y = element_blank(),
            strip.background =element_rect(fill="white"))
    #plot.v

#### FC plot: TREATMENT ####
dat.t <- FC.plot.dat %>% 
  filter(group2 != "RV-infected vs media" &
           term.ord %in% c("INTERFERON\nALPHA\nRESPONSE",
                           "INTERFERON\nGAMMA\nRESPONSE",
                           "INFLAMMATORY\nRESPONSE",
                           "EPITHELIAL\nMESENCHYMAL\nTRANSITION"))

plot.t <- dat.t %>% 
  ggplot(aes(x = group1, y=value)) +
  geom_violin() +
  #Add non-labeled points
  geom_jitter(data=filter(dat.t, is.na(to.label)),
              aes(color=col.group), position = pos) +
  #Add labeled points
  geom_point(data=filter(dat.t, !is.na(to.label)),
             aes(color=col.group), ) +
  #Add labels left
  geom_text_repel(data=filter(dat.t, !is.na(to.label)),
                  aes(label=hgnc_symbol), direction="y",
                  hjust=1, nudge_x=-0.4,
                  show.legend = FALSE, size=3) +
  #Add line at 0
  #geom_hline(yintercept = 0, size=0.5) +
  #Add mean lines
  stat_summary(fun="mean", geom="crossbar")+
  facet_grid(term.ord~group2, scales="free") +
  coord_flip() +
  #Beautify
  theme_bw() +
  labs(x="", y="Gene mean log2 fold change") +
  scale_color_manual(values=c("down"="#6baed6","up"="#ef3b2c")) +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background =element_rect(fill="white"))
#plot.t

#### Combine FC plots ####
fc.plot.all <- plot_grid(plot.v, plot.t, ncol=2, rel_widths = c(0.6,1))
fc.plot.all
ggsave("publication/fig/H_GSEA_FoldChange.pdf", fc.plot.all, 
       width = 20, height = 15)
```

## Genes of interest

```{r}
# List all genes to plot
GOI <- c("CXCL10", "CXCL11")

#Get results for genes of interest
dat.GOI <- bind_rows(h.combo.plot_1, h.combo.plot_2) %>% 
  filter(hgnc_symbol %in% GOI) %>% 
  dplyr::select(-term) %>% distinct()

plot.GOI <- dat.GOI %>% 
  #Beautify labels
  mutate(x.lab=paste(virus.detail,IL5, sep="_"),
         x.lab=gsub("oldH|newH", "", x.lab),
         x.lab=recode_factor(factor(x.lab),
                             "none_none"="media",
                             "none_EOS.supp"="+ EOS sup",
                             "none_AntiIL5"="+ Anti-IL5",
                             "RV_none"="+ RV",
                             "RV_EOS.supp"="+ RV\n+ EOS sup",
                             "RV_AntiIL5"="+ RV\n+ Anti-IL5")) %>% 
  mutate(ex.lab = recode_factor(factor(experiment),
                      "P259.1"="EOS supernatant\nin healthy donors",
                      "P259.2"="Anti-IL5 therapy\nin asthma donors")) %>% 
  arrange(x.lab) %>%

  #Plot
  ggplot(aes(x=x.lab, y=expression, color=donorID)) +
  geom_jitter(width=0.1, height=0) +
  #Add mean and error bars
  stat_summary(fun.data=mean_sdl, 
               fun.args = list(mult=1), 
               geom="errorbar", color="black", width=0.1) +
  stat_summary(fun=mean, geom="errorbar", 
               aes(ymax=..y.., ymin=..y..),
               color="black", width=0.25) +
  #facet experiments and genes
  facet_grid(hgnc_symbol~ex.lab, scales="free") +
  #Beautify
  theme_bw() +
  labs(x="", y="Normalized log2 expression") +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background =element_rect(fill="white"))

plot.GOI
ggsave("publication/fig/GOI.pdf", plot.GOI,
       height=6, width=6)
```

# R session

```{r}
sessionInfo()
```

***

### 2021-05-19 3.P259.pDC_GSEA.Rmd
Update Broad gene sets to v7.4

@@
 GSEA(gene_list = gene.ls, 
-     gmt_file="data_clean/Broad_gmt/h.all.v7.1.symbols.gmt")
+     gmt_file="data_clean/Broad_gmt/h.all.v7.4.symbols.gmt")
 ```
 
-#### Curated gene sets (C2)
+#### Curated canonical pathway gene sets (C2:CP)
 
@@
 GSEA(gene_list = gene.ls, 
-     gmt_file="data_clean/Broad_gmt/c2.all.v7.1.symbols.gmt")
+     gmt_file="data_clean/Broad_gmt/c2.cp.v7.4.symbols.gmt")
 ```
 
-#### GO gene sets (C5)
+#### GO biological procress gene sets (C5:GO:BP)
 
@@
 GSEA(gene_list = gene.ls, 
-     gmt_file="data_clean/Broad_gmt/c5.all.v7.1.symbols.gmt")
+     gmt_file="data_clean/Broad_gmt/c5.go.bp.v7.4.symbols.gmt")
 ```
@@
 ```{r echo=FALSE, message=FALSE}
+fdr.cut <- 0.1
 #Load results for faster knitting
-h_GSEA.result <- read_csv("results/GSEA_FoldChange/h_GSEA.result.csv")
-c2_GSEA.result <- read_csv("results/GSEA_FoldChange/c2_GSEA.result.csv")
-c5_GSEA.result <- read_csv("results/GSEA_FoldChange/c5_GSEA.result.csv")
+h_GSEA.result <- read_csv("results/GSEA/h_GSEA.result.csv")
+c2_GSEA.result <- read_csv("results/GSEA/c2_GSEA.result.csv")
+c5_GSEA.result <- read_csv("results/GSEA/c5_GSEA.result.csv")
+
+#Combine and filter signif
+GSEA.all <- bind_rows(h_GSEA.result,c2_GSEA.result,c5_GSEA.result)
+#list all terms with at least 1 significant treatment contrast
+term.OI.t <- GSEA.all %>% 
+  filter(fgsea.FDR <= fdr.cut & gage.FDR <= fdr.cut) %>% 
+  filter(group %in% c("AntiIL5_nonenone_none", "AntiIL5_HRVnone_HRV",
+                      "EOSsupp_nonenone_none", "EOSsupp_HRVnone_HRV")) %>% 
+  distinct(pathway) %>% unlist(use.names=FALSE)
+
+#list all terms with at least 1 significant virus contrast 
+term.OI.v <- GSEA.all %>% 
+  filter(fgsea.FDR <= fdr.cut & gage.FDR <= fdr.cut) %>% 
+  filter(group %in% c("none_HRVnone_none","AntiIL5_HRVAntiIL5_none",
+                      "EOSsupp_HRVEOSsupp_none")) %>% 
+  distinct(pathway) %>% unlist(use.names=FALSE)
+
+#Find terms with both t and v cutoffs
+term.OI <- intersect(term.OI.t, term.OI.v)
+
+GSEA.signif <- GSEA.all %>% 
+  filter(pathway %in% term.OI)
 ```
@@
 Results are only considered significant if both fgsea and gage methods meet the FDR threshold in the same fold change direction.
+
+```{R echo=FALSE}
+GSEA.signif %>% 
+  select(pathway, group, fgsea.FDR,fgsea.NES) %>% 
+  separate(pathway, into=c("set","pathway"), sep="_", extra = "merge") %>% 
+  select(set, pathway, group, fgsea.FDR,fgsea.NES) %>% 
+  arrange(set, pathway, fgsea.NES) %>% 
+  kable() %>%
+kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
+  collapse_rows(1:2, valign = "top")
+```
 

### 2021-05-19 3.P259.pDC_GSEA.Rmd
Add DEG enrichment analysis

@@
 
+# Hypergeometric enrichment of DEGs
+
 ```{r echo=FALSE}
-source("scripts/GSEA.score.plot.R")
+fdr.cut <- 0.1
 ```
 
-#### Hallmark FDR < 0.1
+Test for Broad gene set enrichment in differentially expressed genes (DEG) as defined by change with virus (*e.g.* significant for virus in untreated and/or treated donors) AND different between untreated and treated donors in media OR RV.
+DEGs are defined at FDR < `r fdr.cut`
 
-```{r echo=FALSE, fig.height=5, fig.width=8.5}
-gsea.plot(gsea = h_GSEA.result, fdr.cut = 0.1, 
-          prefix = "H", print.plot = TRUE, save.plot = FALSE)
+```{r}
+#Define change with virus
+DEG1.v <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
+  filter(group %in% c("none_HRV - none_none", "EOS.supp_HRV - EOS.supp_none") &
+           adj.P.Val <= fdr.cut)
+#Define change with treament
+DEG1.t <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
+  filter(group %in% c("EOS.supp_none - none_none", "EOS.supp_HRV - none_HRV") &
+           adj.P.Val <= fdr.cut)
+
+#Define change with virus
+DEG2.v <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
+  filter(group %in% c("none_HRV - none_none", "AntiIL5_HRV - AntiIL5_none") &
+           adj.P.Val <= fdr.cut)
+#Define change with treament
+DEG2.t <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
+  filter(group %in% c("AntiIL5_none - none_none", "AntiIL5_HRV - none_HRV") &
+           adj.P.Val <= fdr.cut)
+
+#Format to ENTREZ
+
+#Format in list
+DEG.ls <- list()
+DEG.ls[["EOSsup"]] <- intersect(DEG1.v$hgnc_symbol, DEG1.t$hgnc_symbol)
+DEG.ls[["AntiIL5"]] <- intersect(DEG2.v$hgnc_symbol, DEG2.t$hgnc_symbol)
 ```
 
-#### Curated gene sets (C2) FDR < 0.05
-
-```{r echo=FALSE, fig.height=25, fig.width=8.5}
-gsea.plot(gsea = c2_GSEA.result, fdr.cut = 0.05, 
-          prefix = "C2", print.plot = TRUE, save.plot = FALSE)
+```{r message=FALSE, warning=FALSE}
+#Script for running term enrichment
+source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/hypergeo_enricher.R")
 ```
 
-#### GO gene sets (C5) FDR < 0.05
+## Run enrichment
+#### Hallmark (H)
 
-```{r echo=FALSE, fig.height=25, fig.width=8.5}
-gsea.plot(gsea = c5_GSEA.result, fdr.cut = 0.05, 
-          prefix = "C5", print.plot = TRUE, save.plot = FALSE)
+```{r h.2, warnings=FALSE, message=FALSE, eval=FALSE}
+enrich.fxn(gene.list = DEG.ls, ID.type="SYMBOL",
+           category = "H",
+           genome = "org.Hs.eg.db", 
+           basename = "DEG",
+           outdir = "results/enrichment/")
+```
+
+#### Curated canonical pathway gene sets (C2:CP)
+
+```{r c2.2, warnings=FALSE, message=FALSE, eval=FALSE}
+enrich.fxn(gene.list = DEG.ls, ID.type="SYMBOL",
+           category = "C2", subcategory="CP",
+           genome = "org.Hs.eg.db", 
+           basename = "DEG",
+           outdir = "results/enrichment/")
+```
+
+#### GO biological procress gene sets (C5:GO:BP)
+
+```{r c5.2, warnings=FALSE, message=FALSE, eval=FALSE}
+enrich.fxn(gene.list = DEG.ls, ID.type="SYMBOL",
+           category = "C5", subcategory = "GO:BP",
+           genome = "org.Hs.eg.db", 
+           basename = "DEG",
+           outdir = "results/enrichment/")
+```
+
+## Significant enrichment
+
+```{r echo=FALSE, message=FALSE}
+fdr.cut <- 0.05
+#Load results for faster knitting
+h_enrich.result <- read_csv("results/enrichment/enrich_DEG_H.csv")
+c2_enrich.result <- read_csv("results/enrichment/enrich_DEG_C2_CP.csv")
+c5_enrich.result <- read_csv("results/enrichment/enrich_DEG_C5_GO.BP.csv")
+
+#Combine and filter signif
+enrich.signif <- bind_rows(h_enrich.result,c2_enrich.result,c5_enrich.result)%>% 
+  filter(p.adjust <= fdr.cut) 
+```
+
+Gene sets enriched in DEGs at FDR < `r fdr.cut`
+
+```{R echo=FALSE}
+enrich.signif %>% 
+  separate(Description, into=c("set","pathway"), sep="_", extra = "merge") %>% 
+  dplyr::select(group, set, pathway, p.adjust, SYMBOLs) %>% 
+  arrange(group, set, p.adjust) %>% 
+  kable() %>%
+kable_styling(bootstrap_options = "striped", full_width = FALSE) %>% 
+  collapse_rows(1:2, valign = "top")
 ```

### 2021-05-19 publication/scripts/Fig5.PCA.R
Mv PCA to supplemental and add viral batch plot

@@ -18,6 +18,25 @@ group.cols <- c("none:none"="#dadaeb",
 load("data_clean/P259_pDC_clean.RData")
 
 #### Calculate PCA ####
+## All
+samps0 <- dat.pDC.voom$targets %>% 
+  filter(virus=="HRV" & experiment == "P259_2")
+PCA0 <- as.data.frame(dat.pDC.voom_1$E) %>% 
+  rownames_to_column() %>% 
+  full_join(rownames_to_column(as.data.frame(dat.pDC.voom_2$E))) %>% 
+  column_to_rownames() %>% 
+  select(all_of(samps0$libID)) %>% 
+  t() %>% 
+  prcomp(scale. = TRUE)
+
+##Axes labels
+PC1.label0 <- paste("PC1 (", 
+                    round(summary(PCA0)$importance[2,1]*100, digits=1), 
+                    "%)", sep="")
+PC2.label0 <-paste("PC2 (", 
+                   round(summary(PCA0)$importance[2,2]*100, digits=1), 
+                   "%)", sep="")
+
 ## EOS supernatant experiment
 PCA1 <- as.data.frame(dat.pDC.voom_1$E) %>% 
   t() %>% 
@@ -45,11 +64,13 @@ PC2.label2 <-paste("PC2 (",
                    "%)", sep="")
 
 # Extract PC values and merge
-PCA.dat <- as.data.frame(PCA1$x) %>% 
+PCA.dat <- as.data.frame(PCA0$x) %>% 
   rownames_to_column("libID") %>%
+  mutate(calc="all") %>% 
+  bind_rows(rownames_to_column(as.data.frame(PCA1$x),"libID")) %>% 
   bind_rows(rownames_to_column(as.data.frame(PCA2$x),"libID")) %>% 
   # Select PCs
-  dplyr::select(libID, PC1, PC2) %>% 
+  dplyr::select(libID, PC1, PC2, calc) %>% 
   # Merge with metadata
   left_join(dat.pDC.voom$targets, by="libID") %>% 
   #Add facet labels
@@ -59,8 +80,24 @@ PCA.dat <- as.data.frame(PCA1$x) %>%
                             "P259_2" = 'italic("In vivo")~"Anti-IL-5/5R"*alpha'))
 
 #### Plots ####
-PCA1 <- PCA.dat %>% 
-  filter(experiment=="P259_1") %>% 
+PCA.dat0 <- PCA.dat %>% 
[…]

### 2021-05-19 publication/scripts/Fig6.GSEA.R
Add y facet labels to GSEA plot

@@ -5,50 +5,58 @@ library(limma)
 set.seed(4389)
 
 #### Data ####
-H <- read_csv("results/GSEA_FoldChange/h_GSEA.result.csv")
+H <- read_csv("results/GSEA/h_GSEA.result.csv")
 fdr.cut <- 0.1
 
 #### Subset significant ####
 #list all terms with at least 1 significant treatment contrast
 term.OI.t <- H %>% 
   filter(fgsea.FDR <= fdr.cut & gage.FDR <= fdr.cut) %>% 
-  filter(group %in% c("AntiIL5_none.none_none.2", "AntiIL5_HRV.none_HRV.2",
-                      "EOS.supp_none.none_none.1", "EOS.supp_HRV.none_HRV.1")) %>% 
+  filter(group %in% c("AntiIL5_nonenone_none", "AntiIL5_HRVnone_HRV",
+                      "EOSsupp_nonenone_none", "EOSsupp_HRVnone_HRV")) %>% 
   distinct(pathway) %>% unlist(use.names=FALSE)
 
 #list all terms with at least 1 significant virus contrast 
 term.OI.v <- H %>% 
   filter(fgsea.FDR <= fdr.cut & gage.FDR <= fdr.cut) %>% 
-  filter(group %in% c("none_HRV.none_none.2","AntiIL5_HRV.AntiIL5_none.2",
-                      "none_HRV.none_none.1","EOS.supp_HRV.EOS.supp_none.1")) %>% 
+  filter(group %in% c("none_HRVnone_none","AntiIL5_HRVAntiIL5_none",
+                      "EOSsupp_HRVEOSsupp_none")) %>% 
   distinct(pathway) %>% unlist(use.names=FALSE)
 
 #Find terms with both t and v cutoffs
 term.OI <- intersect(term.OI.t, term.OI.v)
 
+
 #### Combine data and format labels ####
 h.plot.dat <- H %>% 
   filter(pathway %in% term.OI) %>% 
-  mutate(group1 = recode_factor(factor(group),
-                                "AntiIL5_HRV.AntiIL5_none.2"='"+ Anti-IL-5/5R"*alpha',
-                                "none_HRV.none_none.2"='"- Anti-IL-5/5R"*alpha',
-                                "EOS.supp_HRV.EOS.supp_none.1"='"+ EOS sup"',
-                                "none_HRV.none_none.1"='"- EOS sup"',
-                                
-                                "EOS.supp_HRV.none_HRV.1"="+ RV",
[…]

### 2021-05-19 publication/scripts/Fig7.GSEA.fold.change.R
Explore leading edge gene labels in GSEA fold change plot

@@ -7,7 +7,7 @@ set.seed(4389)
 
 #### Data ####
 #List genes in each Hallmark term of interest
-myGO <- fgsea::gmtPathways("data_clean/Broad_gmt/h.all.v7.1.symbols.gmt")
+myGO <- fgsea::gmtPathways("data_clean/Broad_gmt/h.all.v7.4.symbols.gmt")
 GO.df <- plyr::ldply(myGO, rbind) %>% 
   rename(term = `.id`) %>% 
   pivot_longer(-term, values_to="hgnc_symbol")
@@ -100,7 +100,7 @@ h.combo.plot <- bind_rows(h.combo.plot_1, h.combo.plot_2) %>%
 #Set jitter
 pos <- position_jitter(width = 0.3, seed = 589, height = 0)
 
-#### List genes to label ####
+#### List top genes to label ####
 #Top N genes within panel
 no.genes <- 2
 #Significant Hallmark terms (in Fig 6)
@@ -190,6 +190,102 @@ to.label.df <- data.table::rbindlist(to.label, idcol = "term") %>%
 h.combo.plot.lab <- h.combo.plot %>% 
   left_join(to.label.df, by = c("term", "hgnc_symbol"))
 
+#### List leading edge genes to label ####
+gsea.LE <- read_csv("results/GSEA/h_GSEA.result.csv") %>% 
+  filter(pathway %in% c("HALLMARK_INTERFERON_ALPHA_RESPONSE",
+                        "HALLMARK_INTERFERON_GAMMA_RESPONSE",
+                        "HALLMARK_INFLAMMATORY_RESPONSE",
+                        "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION")) %>% 
+  #Add experi group
+  mutate(experi = c(rep("AntiIL5",nrow(.)/2),rep("EOSsup",nrow(.)/2))) %>% 
+  select(experi, group, pathway, fgsea.leadingEdge) %>% 
+  #Add FC group
+  mutate(group2 = ifelse(experi=="AntiIL5" & 
+                           group %in% c("none_HRVnone_none", "AntiIL5_HRVAntiIL5_none"), 
+                         "AntiIL5.RV",
+                         ifelse(experi=="EOSsup" & 
+                                  group %in% c("none_HRVnone_none", "EOSsupp_HRVEOSsupp_none"), 
+                                "EOSsup.RV",
[…]

### 2021-05-21 publication/scripts/Fig6.GSEA.R
Clean GSEA facet labels

@@ -65,14 +65,13 @@ h.plot.dat <- H %>%
                                         gage.FDR <= 0.1 & 
                                         fgsea.FC == gage.FC, 
                                       "FDR < 0.1", "NS"))) %>% 
-  mutate(pathway = gsub("HALLMARK_", "", pathway),
-         pathway2 = gsub("_", "\n", pathway)) %>% 
+  mutate(pathway = gsub("HALLMARK_", "", pathway)) %>% 
   #Reorder terms
-  mutate(pathway.ord = factor(pathway2,
-                              levels=c("INTERFERON\nALPHA\nRESPONSE",
-                              "INTERFERON\nGAMMA\nRESPONSE",
-                              "INFLAMMATORY\nRESPONSE",
-                              "EPITHELIAL\nMESENCHYMAL\nTRANSITION")))
+  mutate(pathway.ord = recode_factor(factor(pathway),
+                                     "INTERFERON_ALPHA_RESPONSE"="IFNA response",
+                                     "INTERFERON_GAMMA_RESPONSE"="IFNG response",
+                                     "INFLAMMATORY_RESPONSE"="Inflammatory\nresponse",
+                                     "EPITHELIAL_MESENCHYMAL_TRANSITION"="Epithelial\nmesenchymal\ntransition"))
 
 #### Plot A: virus ####
 h.plot1 <- h.plot.dat %>% 

### 2021-05-21 publication/scripts/Fig7.GSEA.fold.change.R

@@ -90,8 +90,7 @@ h.combo.plot <- bind_rows(h.combo.plot_1, h.combo.plot_2) %>%
   summarise(value = mean(value, na.rm=TRUE)) %>% 
   ungroup() %>% 
   #Beautify for plotting
-  mutate(term = gsub("HALLMARK_","", term),
-         term = gsub("_"," ", term)) %>% 
+  mutate(term = gsub("HALLMARK_","", term)) %>% 
   mutate(col.group = ifelse(value <0, "down", "up")) %>% 
   mutate(group = factor(group, levels=c("treat.media","treat.virus",
                                         "virus.untreat","virus.treat")))
@@ -101,94 +100,94 @@ h.combo.plot <- bind_rows(h.combo.plot_1, h.combo.plot_2) %>%
 pos <- position_jitter(width = 0.3, seed = 589, height = 0)
 
 #### List top genes to label ####
-#Top N genes within panel
-no.genes <- 2
-#Significant Hallmark terms (in Fig 6)
-term.OI.ls <- c("INTERFERON ALPHA RESPONSE", "INTERFERON GAMMA RESPONSE",
-                "INFLAMMATORY RESPONSE",
-                "EPITHELIAL MESENCHYMAL TRANSITION")
-
-to.label <- list()
-
-for(term.OI in term.OI.ls){
-  #List genes in congruent direction
-  #Virus
-  if(term.OI == "EPITHELIAL MESENCHYMAL TRANSITION"){
-    genes.v <- h.combo.plot %>% 
-      #virus fold changes
-      filter(term ==term.OI &
-               group %in% c("virus.treat","virus.untreat")) %>% 
-      #All values negative (fold change down)
-      group_by(term, hgnc_symbol) %>% 
-      filter(max(value, na.rm=TRUE)<0) %>% 
-      ungroup() %>% 
-      distinct(hgnc_symbol) %>% unlist(use.names = FALSE)
-  } else{
-    genes.v <- h.combo.plot %>% 
-      #virus fold changes
-      filter(term ==term.OI &
-               group %in% c("virus.treat","virus.untreat")) %>% 
-      #All values position (fold change up)
-      group_by(term, hgnc_symbol) %>% 
-      filter(min(value, na.rm=TRUE)>0) %>% 
-      ungroup() %>% 
-      distinct(hgnc_symbol) %>% unlist(use.names = FALSE)
-  }
-  
[…]

### 2021-05-21 publication/scripts/FigSX.STRING.R
Add STRING network

@@ -1,0 +1,80 @@
+library(tidyverse)
+library(biomaRt)
+source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/STRING_network_fxn.R")
+set.seed(8434)
+
+#### DEG ####
+fdr.cut <- 0.1
+#Define change with virus
+DEG1.v <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
+  filter(group %in% c("none_HRV - none_none", "EOS.supp_HRV - EOS.supp_none") &
+           adj.P.Val <= fdr.cut)
+#Define change with treament
+DEG1.t <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
+  filter(group %in% c("EOS.supp_none - none_none", "EOS.supp_HRV - none_HRV") &
+           adj.P.Val <= fdr.cut)
+
+#Define change with virus
+DEG2.v <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
+  filter(group %in% c("none_HRV - none_none", "AntiIL5_HRV - AntiIL5_none") &
+           adj.P.Val <= fdr.cut)
+#Define change with treament
+DEG2.t <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
+  filter(group %in% c("AntiIL5_none - none_none", "AntiIL5_HRV - none_HRV") &
+           adj.P.Val <= fdr.cut)
+
+#Format in list
+DEG <- c(intersect(DEG1.v$hgnc_symbol, DEG1.t$hgnc_symbol),
+         intersect(DEG2.v$hgnc_symbol, DEG2.t$hgnc_symbol))
+##replace HGNC symbol with one in STRING
+DEG <- gsub("MARCHF2","MARCH2",DEG)
+DEG <- gsub("H2BC21","HIST2H2BE",DEG)
+DEG <- gsub("H4C14","HIST2H4A",DEG)
+
+#### Enrichment ####
+H <- read_csv("results/enrichment/enrich_DEG_H.csv") %>% 
+  filter(p.adjust <= 0.05 & size.overlap.term >= 2)
+C2 <- read_csv("results/enrichment/enrich_DEG_C2_CP.csv") %>% 
+  filter(p.adjust <= 0.05 & size.overlap.term >= 2)
+
+enrich <- bind_rows(H,C2) %>% 
+  mutate(Description = recode_factor(factor(Description), 
[…]

### 2021-05-21 publication/scripts/Fig8.genes.of.interest.R
Combine DEG boxplots and STRING

@@ -9,10 +9,11 @@ load("data_clean/P259_pDC_clean.RData")
 
 #### List all genes to plot ####
 #Genes to label from Fig7
-load("publication/fig/to.label.RData")
+load("publication/fig/to.label.DEG.RData")
+load("publication/fig/to.label.LE.RData")
 
-## IFN and inflam. None for epiMT
-GOI <- unique(unlist(to.label[1:3], use.names=FALSE))
+LE <- unique(unlist(to.label[1:3], use.names=FALSE))
+GOI <- intersect(c(DEG1,DEG2),LE)
 #Add IFN genes
 GOI2 <- c("IFNA2","IFNG")
 
@@ -142,8 +143,8 @@ GOI.pey <- GOI.pe %>%
   ungroup()
 
 #### Plot GSEA genes ####
-plot.GOI1 <- dat.GOI.anno %>% 
-  filter(hgnc_symbol %in% GOI) %>% 
+plot.GOI1a <- dat.GOI.anno %>% 
+  filter(hgnc_symbol %in% sort(GOI)[1:4]) %>% 
   droplevels() %>% 
   
   ggplot(aes(x=x.lab, y=2^expression, color=donorID)) +
@@ -157,7 +158,7 @@ plot.GOI1 <- dat.GOI.anno %>%
   facet_grid(hgnc_symbol~facet.lab, scales="free", 
              labeller = labeller(facet.lab=label_parsed)) +
   #Add pval
-  stat_pvalue_manual(data=filter(GOI.pey, hgnc_symbol %in% GOI), 
+  stat_pvalue_manual(data=filter(GOI.pey, hgnc_symbol %in% sort(GOI)[1:4]), 
                      label="symbol", xmin="group1", xmax="group2") +
   # #Beautify
   theme_bw() +
@@ -170,7 +171,36 @@ plot.GOI1 <- dat.GOI.anno %>%
         plot.margin = margin(0.2,0.2,0.2,0.2,"cm")) +
   scale_x_discrete(labels = ggplot2:::parse_safe)
 
-#plot.GOI1
+plot.GOI1b <- dat.GOI.anno %>% 
+  filter(hgnc_symbol %in% sort(GOI)[5:8]) %>% 
+  droplevels() %>% 
+  
+  ggplot(aes(x=x.lab, y=2^expression, color=donorID)) +
+  geom_jitter(width=0.1, height=0) +
+  stat_summary(fun.data=mean_sdl, 
+               fun.args = list(mult=1), 
[…]

### 2021-05-21 publication/scripts/Table.formatting.R
Mv publication table formatting to script

@@ -1,0 +1,78 @@
+#Create directories of outputs
+dir.create("publication/table/", showWarnings = FALSE, recursive = TRUE)
+
+#Load packages used across all data
+library(tidyverse)
+library(limma)
+#Set seed
+set.seed(4389)
+
+
+# Tables
+## RNA-seq library and patient metadata
+attach("data_clean/P259_pDC_clean.RData")
+
+dat.pDC.voom$targets %>% 
+  select(libID, everything()) %>% 
+  arrange(libID) %>% 
+  write_csv(., "publication/table/TableS.metadata.csv")
+
+### Patient summary demographics
+#Age
+dat.pDC.voom$targets %>% 
+  distinct(experiment, donorID, Age) %>% 
+  group_by(experiment) %>% 
+  summarise(.groups = "keep",
+            n = n(),
+            mean.age = mean(Age, na.rm=TRUE),
+            sd.age = sd(Age, na.rm=TRUE)/sqrt(n))
+#Sex
+dat.pDC.voom$targets %>% 
+  distinct(experiment, donorID, Sex) %>% 
+  count(experiment, Sex) %>% 
+  group_by(experiment) %>% 
+  mutate(pct = n/sum(n))
+
+## RNA-seq normalized log2 counts
+as.data.frame(dat.pDC.voom$E) %>% 
+  rownames_to_column("geneName") %>% 
+  #Save
+  write_csv(., "publication/table/TableS.norm.log2.counts.csv")
+
+## All contrast model results
+#Get all model results
+read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
+  #Add experiment variable
+  mutate(experiment = "P259_1") %>% 
+  bind_rows(read_csv("results/gene_level/P259.2_gene_pval.csv")) %>% 
+  #Fill in experiment variable
+  mutate(experiment = ifelse(is.na(experiment), "P259_2", experiment)) %>% 
+  #Keep only contrasts model 
+  filter(model == "contrasts") %>% 
+  #Reorder and rename variables
+  select(experiment, model, geneName, hgnc_symbol, group,
+         AveExpr:group) %>% 
+  rename(contrast = group) %>% 
+  arrange(experiment, geneName, contrast) %>% 
+  #Save
+  write_csv(., "publication/table/TableS.models.csv")
+
+## GSEA results
+read_csv("results/GSEA_FoldChange/h_GSEA.result.csv") %>% 
+  rename(contrast = group) %>% 
[…]

### 2022-03-18 3.P259.pDC_response.to.review.Rmd
Add response to JACI review and update figures

---
title: 'P259: Response to reviews'
subtitle: "Dendritic cells (pDC)"
author: "Kim Dill-McFarland, kadm@uw.edu"
date: "version `r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    toc: yes
    toc_depth: '3'
  html_document:
    toc: yes
    toc_depth: 3
    toc_float:
      collapsed: no
editor_options:
  chunk_output_type: console
---
# Background

Analyses performed in response to reviews for publication in JACI.

# Setup
Load packages

```{r message=FALSE, warning=FALSE}
# Data manipulation and figures
library(tidyverse)
library(ggpubr)
library(readxl)
library(patchwork)
#Print pretty tables to Rmd
library(knitr)
library(kableExtra)
```

Set seed

```{r}
set.seed(589)
```

Custom functions

```{r}
#To extract and format p-value from lmFit
source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/limma.extract.pval.R")
```

# Low-quality libraries

Determine if low-quality libraries removed from analysis are associated with low input RNA. All filtered libraries with low raw sequences and high CV coverage also had low input RNA.

```{r echo=FALSE, message=FALSE, fig.width=8.5}
#### Data ####
attach("data_clean/P259_pDC_clean.RData")
meta <- read_csv("data_clean/P259_pDC_metadata.csv")
anno <- read_excel("data_raw/P259-2 Final Annotation.xlsx")

dat <- meta %>% 
  select(libID, total_sequences, median_cv_coverage) %>% 
  inner_join(select(anno, "library sampleId", "RNA conc (ng/ul)"), 
             by=c("libID"="library sampleId")) %>% 
  mutate(`Quality filter` = ifelse(libID %in% dat.pDC.voom$targets$libID, "Pass", "Fail"))

#### Plot ####
plot1 <- dat %>% 
  ggplot(aes(y=`RNA conc (ng/ul)`, x=total_sequences)) +
  geom_point(aes(color=`Quality filter`), size=2) +
  theme_classic() +
  labs(x = "Raw sequences", y="RNA concentration (ng/ul)")

plot2 <- dat %>% 
  ggplot(aes(y=`RNA conc (ng/ul)`, x=median_cv_coverage)) +
  geom_point(aes(color=`Quality filter`), size=2) +
  theme_classic() +
  labs(x = "Raw sequences", y="RNA concentration (ng/ul)")

plot1+plot2
```

# Viral load
### Data

RV sequences were extracted from RNA-seq libraries and quantified using methods similar to human data. Normalized RV counts were calculated as total RV sequences / total non-human sequences * 1E6

```{r echo=FALSE, message=FALSE}
dat2 <- read_csv("data_raw/211102_P259-1_P259-2_RhinoViruses_normlized_read_counts.csv") %>% 
  rename(libID=libid) %>% 
  #filter libraries in final analysis
  filter(libID %in% dat.pDC.voom$targets$libID) %>% 
  #add metadata
  left_join(select(meta, libID, experiment, donorID, IL5, virus, virus.detail)) %>% 
  mutate(contrast=paste(IL5,virus, sep="_")) %>% 
  mutate(contrast=factor(contrast, 
                         levels = c("none_none","none_HRV",
                                 "AntiIL5_none","AntiIL5_HRV",
                                 "EOS.supp_none","EOS.supp_HRV")))
```

```{r echo=FALSE, message=FALSE}
# Format for limma
count1 <- dat2 %>% 
  filter(experiment == "P259_1") %>% 
  select(libID, RhinoVirusA_normCount) %>% 
  pivot_longer(-libID) %>% 
  arrange(libID) %>% 
  pivot_wider(names_from = libID) %>% 
  column_to_rownames("name")

meta1 <- dat2 %>% 
  filter(experiment == "P259_1") %>% 
  arrange(libID) %>% 
  droplevels()

count2 <- dat2 %>% 
  filter(experiment == "P259_2") %>% 
  select(libID, RhinoVirusA_normCount) %>% 
  pivot_longer(-libID) %>% 
  arrange(libID) %>% 
  pivot_wider(names_from = libID) %>% 
  column_to_rownames("name")

meta2 <- dat2 %>% 
  filter(experiment == "P259_2") %>% 
  arrange(libID) %>% 
  droplevels()
```

### Run limma contrasts model

In both experiments, viral loads were significantly higher in virus-infected samples and there were no differences between EOS supernatant or AntiIL5 treatment groups.

```{r echo=FALSE}
# Define model
model_1.contrast<- model.matrix(~ 0 + contrast, data=meta1)
    colnames(model_1.contrast) <- c(
                                 "none_none","none_HRV",
                                 "EOS.supp_none","EOS.supp_HRV")
#Block by donor  
consensus.corr1 <- duplicateCorrelation(
  count1,
  model_1.contrast,
  block=meta1$donorID)$consensus.correlation

#Fit model
fitQW_1.contrast <- lmFit(count1, 
                      model_1.contrast,
                      block=meta1$donorID,
                      correlation=consensus.corr1)

#Get contrasts
contrast.matrix1 <- makeContrasts(
  none_HRV-none_none, 
  EOS.supp_HRV-EOS.supp_none,
  EOS.supp_none-none_none,
  EOS.supp_HRV-none_HRV,
  levels=model_1.contrast)

efitQW_1.contrast <- eBayes(contrasts.fit(fitQW_1.contrast,
                                  contrast.matrix1))

extract.pval(model=model_1.contrast,
             voom.dat=count1, 
             eFit=efitQW_1.contrast, 
             name="pval_1.contrast",
             summary=FALSE,
             contrasts=TRUE,
             contrast.mat=contrast.matrix1)
```

```{r echo=FALSE}
# Define model
model_2.contrast<- model.matrix(~ 0 + contrast, data=meta2)
    colnames(model_2.contrast) <- c(
                                 "none_none","none_HRV",
                                 "AntiIL5_none","AntiIL5_HRV")
#Block by donor  
consensus.corr2 <- duplicateCorrelation(
  count2,
  model_2.contrast,
  block=meta2$donorID)$consensus.correlation

#Fit model
fitQW_2.contrast <- lmFit(count2, 
                      model_2.contrast,
                      block=meta2$donorID,
                      correlation=consensus.corr2)

#Get contrasts
contrast.matrix2 <- makeContrasts(
  none_HRV-none_none, 
  AntiIL5_HRV-AntiIL5_none,
  AntiIL5_none-none_none,
  AntiIL5_HRV-none_HRV,
  levels=model_2.contrast)

efitQW_2.contrast <- eBayes(contrasts.fit(fitQW_2.contrast,
                                  contrast.matrix2))

extract.pval(model=model_2.contrast,
             voom.dat=count2, 
             eFit=efitQW_2.contrast, 
             name="pval_2.contrast",
             summary=FALSE,
             contrasts=TRUE,
             contrast.mat=contrast.matrix2)
```

```{r echo=FALSE}
pval_1.contrast %>% 
  select(group, logFC, adj.P.Val) %>% 
  kable(align=c("l","c","c"), digits=4, caption="P259_1",
        col.names = c("Variable","Fold change", "FDR")) %>% 
  kable_styling(latex_options="HOLD_position", full_width = FALSE)

pval_2.contrast %>% 
  select(group, logFC, adj.P.Val) %>% 
  kable(align=c("l","c","c"), digits=4, caption="P259_2", 
        col.names = c("Variable","Fold change", "FDR")) %>% 
  kable_styling(latex_options="HOLD_position", full_width = FALSE)
```

```{r echo=FALSE}
#save
write_csv(pval_1.contrast, file="review_response/RV.EOS.model.csv")
write_csv(pval_2.contrast, file="review_response/RV.antiIL5.model.csv")
```

### Plot

```{r echo=FALSE, fig.width=8.5, message=FALSE, warning=FALSE}
#### Format pval data ####
#Pvals for eos experiment
GOI.p1 <- pval_1.contrast %>% 
  mutate(dataset="P259.1")

#Pvals for aniti-IL5 experiment
GOI.p <- pval_2.contrast%>% 
  mutate(dataset="P259.2") %>% 
  #Combine with eos data
  bind_rows(GOI.p1) %>% 
  #Make symbols for plots
  mutate(symbol = ifelse(P.Value <= 0.001,"***",
                         ifelse(P.Value <= 0.01, "**",
                                ifelse(P.Value <= 0.05, "*", NA)))) %>% 
  filter(!is.na(symbol)) %>% 
  #Match group labels to those in plots
  mutate(group.long = paste(dataset, group, sep="_"),
         group1 = recode_factor(factor(group.long),
                        "P259.2_none_HRV - none_none"='"-\n-"',
                        "P259.2_AntiIL5_HRV - AntiIL5_none"='"-\n+"',
                        "P259.2_AntiIL5_none - none_none"='"-\n-"',
                        "P259.2_AntiIL5_HRV - none_HRV"='"+\n-"',
                        "P259.1_none_HRV - none_none"='"RV            -\nEOS sup  -"',
                        "P259.1_EOS.supp_HRV - EOS.supp_none"='"-\n+"',
                        "P259.1_EOS.supp_none - none_none"='"RV            -\nEOS sup  -"',
                        "P259.1_EOS.supp_HRV - none_HRV"='"+\n-"'),
         group2 = recode_factor(factor(group.long),
                          "P259.2_none_HRV - none_none"='"+\n-"',
                          "P259.2_AntiIL5_HRV - AntiIL5_none"='"+  RV\n+  Anti-IL-5/5R"*alpha',
                          "P259.2_AntiIL5_none - none_none"='"-\n+"',
                          "P259.2_AntiIL5_HRV - none_HRV"='"+  RV\n+  Anti-IL-5/5R"*alpha',
                          "P259.1_none_HRV - none_none"='"+\n-"',
                          "P259.1_EOS.supp_HRV - EOS.supp_none"='"+\n+"',
                          "P259.1_EOS.supp_none - none_none"='"-\n+"',
                          "P259.1_EOS.supp_HRV - none_HRV"='"+\n+"'),
         facet.lab = recode_factor(factor(dataset),
                            "P259.1" = '"EOS sup"',
                            "P259.2" = '"Anti-IL-5/5R"*alpha')) %>% 
  dplyr::select(facet.lab, group1, group2, symbol)

#Add y location for pval based on max expression in plot
GOI.pe <- dat2 %>% 
  #Max expression per gene and experiment
  group_by(experiment) %>% 
  summarise(max.e = max(RhinoVirusA_normCount, na.rm=TRUE)) %>% 
  ungroup() %>% 
  mutate(facet.lab = experiment,
         facet.lab = recode_factor(factor(facet.lab),
                            "P259_1" = '"EOS sup"',
                            "P259_2" = '"Anti-IL-5/5R"*alpha')) %>% 
  #Add to pval data
  right_join(GOI.p) %>% 
  arrange(experiment, group1, group2)

#first entry per gene
#Set y position to 1
first <- GOI.pe %>% 
  group_by(experiment) %>% 
  slice(1) %>% 
  mutate(y.position1 = 1)

#Add first position data back and fill in remaining
GOI.pey <- GOI.pe %>% 
  full_join(first) %>%

  #Fill in positions 2 - N
  group_by(experiment, facet.lab) %>% 
  mutate(y.position2 = lag(y.position1)+1,
         y.position3 = lag(y.position2)+1,
         y.position4 = lag(y.position3)+1) %>% 
  #Collapse positions into 1 column
  mutate(y.position = ifelse(!is.na(y.position1),y.position1,
                             ifelse(!is.na(y.position2),y.position2,
                                    ifelse(!is.na(y.position3),y.position3,
                                           ifelse(!is.na(y.position4),y.position4,NA))))) %>% 
  #Scale to max expression
  mutate(y.position = max.e+y.position*5E3) %>% 
  ungroup()
#plot
plot3 <- dat2 %>% 
  mutate(x.lab=paste(experiment, virus.detail, IL5, sep="_"),
         x.lab=gsub("oldH|newH", "", x.lab),
         x.lab=recode_factor(factor(x.lab),
                             "P259_1_none_none"='"RV            -\nEOS sup  -"',
                             "P259_2_none_none"='"-\n-"',
                             "P259_1_none_EOS.supp"='"-\n+"',
                             "P259_2_none_AntiIL5"='"-\n+"',
                             "P259_1_RV_none"='"+\n-"',
                             "P259_2_RV_none"='"+\n-"',
                             "P259_1_RV_EOS.supp"='"+\n+"',
                             "P259_2_RV_AntiIL5"='"+  RV\n+  Anti-IL-5/5R"*alpha')) %>% 
  mutate(facet.lab = experiment,
         facet.lab = recode_factor(factor(facet.lab),
                            "P259_1" = '"EOS sup"',
                            "P259_2" = '"Anti-IL-5/5R"*alpha')) %>% 
  arrange(facet.lab) %>% 
  ggplot(aes(x=x.lab, y=RhinoVirusA_normCount, color=donorID)) +
  geom_jitter(width=0.1, height=0) +
  stat_summary(fun.data=mean_sdl, 
               fun.args = list(mult=1), 
               geom="errorbar", color="black", width=0.25) +
  stat_summary(fun=mean, geom="errorbar", 
               aes(ymax=..y.., ymin=..y..),
               color="black", width=0.5) +
  facet_grid(~facet.lab, scales="free", 
             labeller = labeller(facet.lab=label_parsed)) +
  #Add pval
  stat_pvalue_manual(data=GOI.pey,
                     label="symbol", xmin="group1", xmax="group2") +
  # #Beautify
  theme_bw() +
  labs(x="", y="Normalized RV expression") +
  theme(legend.position = "none",
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        strip.background =element_rect(fill="white"),
        axis.text.x=element_text(hjust=0.95,vjust=-0.5),
        plot.margin = margin(0.2,0.2,0.2,0.2,"cm")) +
  scale_x_discrete(labels = ggplot2:::parse_safe)
plot3
```

# R session

```{r}
sessionInfo()
```

***

### 2022-03-18 publication/scripts/Fig5.GSEA.R

@@ -45,14 +45,14 @@ h.plot.dat <- H %>%
                          ifelse(group1 == "none_HRVnone_none" & experi == "EOSsup",
                                 '"- EOS sup"', group1)),
          group2 = recode_factor(factor(group),
-            "none_HRVnone_none"='italic("Ex vivo")~"RV-infected vs media"',
-            "EOSsupp_HRVEOSsupp_none"='italic("Ex vivo")~"RV-infected vs media"',
-            "AntiIL5_HRVAntiIL5_none"='italic("Ex vivo")~"RV-infected vs media"',
+            "none_HRVnone_none"='"RV-infected vs media"',
+            "EOSsupp_HRVEOSsupp_none"='"RV-infected vs media"',
+            "AntiIL5_HRVAntiIL5_none"='"RV-infected vs media"',
                                 
-            "EOSsupp_HRVnone_HRV"='italic("Ex vivo")~"EOS supernatant vs none"',
-            "AntiIL5_HRVnone_HRV"='italic("In vivo")~"Anti-IL-5/5R"*alpha~"vs none"',
-            "EOSsupp_nonenone_none"='italic("Ex vivo")~"EOS supernatant vs none"',
-            "AntiIL5_nonenone_none"='italic("In vivo")~"Anti-IL-5/5R"*alpha~"vs none"')) %>% 
+            "EOSsupp_HRVnone_HRV"='"EOS sup vs none"',
+            "AntiIL5_HRVnone_HRV"='"Anti-IL-5/5R"*alpha~"vs none"',
+            "EOSsupp_nonenone_none"='"EOS sup vs none"',
+            "AntiIL5_nonenone_none"='"Anti-IL-5/5R"*alpha~"vs none"')) %>% 
   #order factors
   mutate(group1 = factor(group1, levels=c('"+ Anti-IL-5/5R"*alpha','"- Anti-IL-5/5R"*alpha',
                                           '"+ EOS sup"','"- EOS sup"', 
@@ -75,7 +75,7 @@ h.plot.dat <- H %>%
 
 #### Plot A: virus ####
 h.plot1 <- h.plot.dat %>% 
-  filter(group2=='italic("Ex vivo")~"RV-infected vs media"') %>% 
+  filter(group2=='"RV-infected vs media"') %>% 
   
   ggplot(aes(x=group1, y=fgsea.NES)) +
[…]

### 2022-03-18 publication/scripts/Fig6.GSEA.fold.change.R

@@ -1,308 +1,0 @@
-#### Setup ####
-library(tidyverse)
-library(cowplot)
-library(limma)
-library(ggrepel)
-set.seed(4389)
-
-#### Data ####
-#List genes in each Hallmark term of interest
-myGO <- fgsea::gmtPathways("data_clean/Broad_gmt/h.all.v7.4.symbols.gmt")
-GO.df <- plyr::ldply(myGO, rbind) %>% 
-  dplyr::rename(term = `.id`) %>% 
-  pivot_longer(-term, values_to="hgnc_symbol")
-
-#Expression data
-load("data_clean/P259_pDC_clean.RData")
-
-#### Format data ####
-#Select genes of interest and format
-## P259.2
-h.combo.plot_2 <- as.data.frame(dat.pDC.voom_2$E) %>% 
-  rownames_to_column("geneName") %>% 
-  #Get HGNC symbols
-  left_join(dat.pDC.voom_2$genes) %>% 
-  dplyr::select(hgnc_symbol, everything()) %>% 
-  #Filter to genes in terms
-  inner_join(GO.df) %>% 
-  dplyr::select(term, hgnc_symbol, geneName, everything(), 
-                -c(`Previous symbols`:gene_biotype), -name) %>% 
-  #longer format
-  pivot_longer(-c(term:geneName), names_to = "libID", 
-               values_to = "expression") %>% 
-  #Add metadata
-  mutate(experiment = "P259.2") %>% 
-  left_join(dplyr::select(dat.pDC.voom_2$targets, donorID, libID, 
-                          IL5, virus.detail))
-
-## P259.1
-h.combo.plot_1 <- as.data.frame(dat.pDC.voom_1$E) %>% 
-  rownames_to_column("geneName") %>% 
-  #Get HGNC symbols
-  left_join(dat.pDC.voom_1$genes) %>% 
-  dplyr::select(hgnc_symbol, everything()) %>% 
-  #Filter to genes in terms
-  inner_join(GO.df) %>% 
-  dplyr::select(term, hgnc_symbol, geneName, everything(), 
-                -c(`Previous symbols`:gene_biotype), -name) %>% 
-  #longer format
-  pivot_longer(-c(term:geneName), names_to = "libID", 
-               values_to = "expression") %>% 
-  #Add metadata
-  mutate(experiment = "P259.1") %>% 
-  left_join(dplyr::select(dat.pDC.voom_1$targets, donorID,
-                          libID, IL5, virus.detail))
-
[…]

### 2022-03-18 publication/scripts/Fig6.genes.of.interest.R

@@ -1,0 +1,244 @@
+#### Setup ####
+library(tidyverse)
+library(cowplot)
+library(limma)
+set.seed(4389)
+
+#### Data ####
+load("data_clean/P259_pDC_clean.RData")
+
+#### List all genes to plot ####
+#Genes to label from Fig7
+load("publication/fig/to.label.DEG.RData")
+load("publication/fig/to.label.LE.RData")
+
+LE <- unique(unlist(to.label[1:3], use.names=FALSE))
+GOI <- intersect(c(DEG1,DEG2),LE)
+#Add IFN genes
+GOI2 <- c("IFNA2","IFNG")
+
+#### Extract expression data ####
+dat.GOI <- as.data.frame(dat.pDC.voom$E) %>% 
+  rownames_to_column("geneName") %>% 
+  pivot_longer(-geneName, names_to="libID", values_to="expression") %>% 
+  left_join(dat.pDC.voom$targets)
+
+#### Add hgnc symbol ####
+#Get ref genome
+library(biomaRt)
+ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl", 
+                   host = "useast.ensembl.org")
+#Check genome version
+#searchDatasets(mart = ensembl, pattern = "hsapiens")
+
+#Get HGNC symbols
+dat.GOI.anno <- getBM(attributes = c('hgnc_symbol', 'ensembl_gene_id'),
+                      mart = ensembl) %>%
+  #Keep genes in RNAseq data
+  filter(ensembl_gene_id %in% unique(dat.GOI$geneName)) %>% 
+  inner_join(dat.GOI, by=c("ensembl_gene_id"="geneName")) %>% 
+  #Beautify labels
+  mutate(x.lab=paste(experiment, virus.detail,IL5, sep="_"),
+         x.lab=gsub("oldH|newH", "", x.lab),
+         x.lab=recode_factor(factor(x.lab),
+                             "P259_1_none_none"='"RV            -\nEOS sup  -"',
+                             "P259_2_none_none"='"-\n-"',
+                             "P259_1_none_EOS.supp"='"-\n+"',
+                             "P259_2_none_AntiIL5"='"-\n+"',
+                             "P259_1_RV_none"='"+\n-"',
+                             "P259_2_RV_none"='"+\n-"',
+                             "P259_1_RV_EOS.supp"='"+\n+"',
[…]

### 2022-03-18 publication/scripts/Fig7.genes.of.interest.R

@@ -1,243 +1,0 @@
-#### Setup ####
-library(tidyverse)
-library(cowplot)
-library(limma)
-set.seed(4389)
-
-#### Data ####
-load("data_clean/P259_pDC_clean.RData")
-
-#### List all genes to plot ####
-#Genes to label from Fig7
-load("publication/fig/to.label.DEG.RData")
-load("publication/fig/to.label.LE.RData")
-
-LE <- unique(unlist(to.label[1:3], use.names=FALSE))
-GOI <- intersect(c(DEG1,DEG2),LE)
-#Add IFN genes
-GOI2 <- c("IFNA2","IFNG")
-
-#### Extract expression data ####
-dat.GOI <- as.data.frame(dat.pDC.voom$E) %>% 
-  rownames_to_column("geneName") %>% 
-  pivot_longer(-geneName, names_to="libID", values_to="expression") %>% 
-  left_join(dat.pDC.voom$targets)
-
-#### Add hgnc symbol ####
-#Get ref genome
-library(biomaRt)
-ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl", 
-                   host = "useast.ensembl.org")
-#Check genome version
-#searchDatasets(mart = ensembl, pattern = "hsapiens")
-
-#Get HGNC symbols
-dat.GOI.anno <- getBM(attributes = c('hgnc_symbol', 'ensembl_gene_id'),
-                      mart = ensembl) %>%
-  #Keep genes in RNAseq data
-  filter(ensembl_gene_id %in% unique(dat.GOI$geneName)) %>% 
-  inner_join(dat.GOI, by=c("ensembl_gene_id"="geneName")) %>% 
-  #Beautify labels
-  mutate(x.lab=paste(experiment, virus.detail,IL5, sep="_"),
-         x.lab=gsub("oldH|newH", "", x.lab),
-         x.lab=recode_factor(factor(x.lab),
-                             "P259_1_none_none"='"RV            -\nEOS sup  -"',
-                             "P259_2_none_none"='"-\n-"',
-                             "P259_1_none_EOS.supp"='"-\n+"',
-                             "P259_2_none_AntiIL5"='"-\n+"',
-                             "P259_1_RV_none"='"+\n-"',
-                             "P259_2_RV_none"='"+\n-"',
-                             "P259_1_RV_EOS.supp"='"+\n+"',
[…]

### 2022-03-18 publication/scripts/FigE1.PCA.R

@@ -1,137 +1,0 @@
-#### Setup ####
-library(tidyverse)
-library(cowplot)
-library(limma)
-set.seed(4389)
-
-#Define ggplot colors
-group.cols <- c("none:none"="#dadaeb",
-                "none:AntiIL5"="#9e9ac8",
-                "none:EOS.supp"="#54278f",
-                "HRV:none"="#c7e9c0",
-                "HRV:AntiIL5"="#74c476",
-                "HRV:EOS.supp"="#006d2c",
-                "flu:none"="#fdae6b",
-                "flu:AntiIL5"="#e6550d")
-
-#### Data ####
-load("data_clean/P259_pDC_clean.RData")
-
-#### Calculate PCA ####
-## All
-samps0 <- dat.pDC.voom$targets %>% 
-  filter(virus=="HRV" & experiment == "P259_2")
-PCA0 <- as.data.frame(dat.pDC.voom_1$E) %>% 
-  rownames_to_column() %>% 
-  full_join(rownames_to_column(as.data.frame(dat.pDC.voom_2$E))) %>% 
-  column_to_rownames() %>% 
-  select(all_of(samps0$libID)) %>% 
-  t() %>% 
-  prcomp(scale. = TRUE)
-
-##Axes labels
-PC1.label0 <- paste("PC1 (", 
-                    round(summary(PCA0)$importance[2,1]*100, digits=1), 
-                    "%)", sep="")
-PC2.label0 <-paste("PC2 (", 
-                   round(summary(PCA0)$importance[2,2]*100, digits=1), 
-                   "%)", sep="")
-
-## EOS supernatant experiment
-PCA1 <- as.data.frame(dat.pDC.voom_1$E) %>% 
-  t() %>% 
-  prcomp(scale. = TRUE)
-
-##Axes labels
-PC1.label1 <- paste("PC1 (", 
-                    round(summary(PCA1)$importance[2,1]*100, digits=1), 
-                    "%)", sep="")
-PC2.label1 <-paste("PC2 (", 
-                   round(summary(PCA1)$importance[2,2]*100, digits=1), 
-                   "%)", sep="")
-
-## Anti-IL5 therapy experiment
-PCA2 <- as.data.frame(dat.pDC.voom_2$E) %>% 
-  t() %>% 
-  prcomp(scale. = TRUE)
-
-##Axes labels
-PC1.label2 <- paste("PC1 (", 
[…]

### 2022-03-18 publication/scripts/FigE2.GSEA.term.venn.R

@@ -1,41 +1,0 @@
-#### Setup ####
-library(plyr)
-library(tidyverse)
-library(fgsea)
-library(venn)
-
-#### Data ####
-myGO <- fgsea::gmtPathways("data_clean/Broad_gmt/h.all.v7.4.symbols.gmt") %>% 
-  #Convert to data frame
-  plyr::ldply(., data.frame) %>% 
-  dplyr::rename(pathway = `.id`, gene = `X..i..`)
-
-#### Venn ####
-# Get list of genes in pathways
-
-venn.ls <- list()
-
-for(term in c("HALLMARK_INTERFERON_ALPHA_RESPONSE",
-              "HALLMARK_INTERFERON_GAMMA_RESPONSE",
-              "HALLMARK_INFLAMMATORY_RESPONSE",
-              "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION")){
-  
-  #Subset to term of interest
-  myGO.temp <- myGO %>% 
-    filter(pathway == term) %>% 
-    distinct(gene) %>% unlist(use.names = FALSE)
-  
-  #Save to list
-  name <- gsub("HALLMARK_", "", term)
-  name <- gsub("_", "\n", name)
-  
-  venn.ls[[name]] <- myGO.temp
-}
-
-# Plot venn and save
-pdf("publication/fig/FigE2.GSEA.term.venn.pdf",
-    width=5.5, height=5)
-venn(ilab=FALSE, ilcs=1, sncs=1, box = FALSE,
-     x=venn.ls)
-dev.off()
-

### 2022-03-18 publication/scripts/FigE2.PCA.R

@@ -1,0 +1,137 @@
+#### Setup ####
+library(tidyverse)
+library(cowplot)
+library(limma)
+set.seed(4389)
+
+#Define ggplot colors
+group.cols1 <- c("none:none"="#dadaeb",
+                 "none:EOS.supp"="#54278f",
+                 "HRV:none"="#c7e9c0",
+                 "HRV:EOS.supp"="#006d2c")
+group.cols2 <- c("none:none"="#dadaeb",
+                 "none:AntiIL5"="#9e9ac8",
+                 "HRV:none"="#c7e9c0",
+                 "HRV:AntiIL5"="#74c476")
+
+#### Data ####
+load("data_clean/P259_pDC_clean.RData")
+
+#### Calculate PCA ####
+## All
+samps0 <- dat.pDC.voom$targets %>% 
+  filter(virus=="HRV" & experiment == "P259_2")
+PCA0 <- as.data.frame(dat.pDC.voom_1$E) %>% 
+  rownames_to_column() %>% 
+  full_join(rownames_to_column(as.data.frame(dat.pDC.voom_2$E))) %>% 
+  column_to_rownames() %>% 
+  select(all_of(samps0$libID)) %>% 
+  t() %>% 
+  prcomp(scale. = TRUE)
+
+##Axes labels
+PC1.label0 <- paste("PC1 (", 
+                    round(summary(PCA0)$importance[2,1]*100, digits=1), 
+                    "%)", sep="")
+PC2.label0 <-paste("PC2 (", 
+                   round(summary(PCA0)$importance[2,2]*100, digits=1), 
+                   "%)", sep="")
+
+## EOS supernatant experiment
+PCA1 <- as.data.frame(dat.pDC.voom_1$E) %>% 
+  t() %>% 
+  prcomp(scale. = TRUE)
+
+##Axes labels
+PC1.label1 <- paste("PC1 (", 
+                    round(summary(PCA1)$importance[2,1]*100, digits=1), 
+                    "%)", sep="")
+PC2.label1 <-paste("PC2 (", 
+                   round(summary(PCA1)$importance[2,2]*100, digits=1), 
+                   "%)", sep="")
+
+## Anti-IL5 therapy experiment
+PCA2 <- as.data.frame(dat.pDC.voom_2$E) %>% 
+  t() %>% 
+  prcomp(scale. = TRUE)
+
+##Axes labels
+PC1.label2 <- paste("PC1 (", 
[…]

### 2022-03-18 publication/scripts/FigE3.RV.seqs.R

@@ -1,0 +1,147 @@
+library(tidyverse)
+
+#### Data ####
+attach("data_clean/P259_pDC_clean.RData")
+meta <- read_csv("data_clean/P259_pDC_metadata.csv")
+
+dat2 <- read_csv("data_raw/211102_P259-1_P259-2_RhinoViruses_normlized_read_counts.csv") %>% 
+  rename(libID=libid) %>% 
+  #filter libraries in final analysis
+  filter(libID %in% dat.pDC.voom$targets$libID) %>% 
+  #add metadata
+  left_join(select(meta, libID, experiment, donorID, IL5, virus, virus.detail)) %>% 
+  mutate(contrast=paste(IL5,virus, sep="_")) %>% 
+  mutate(contrast=factor(contrast, 
+                         levels = c("none_none","none_HRV",
+                                    "AntiIL5_none","AntiIL5_HRV",
+                                    "EOS.supp_none","EOS.supp_HRV")))
+
+# Format for limma
+count1 <- dat2 %>% 
+  filter(experiment == "P259_1") %>% 
+  select(libID, RhinoVirusA_normCount) %>% 
+  pivot_longer(-libID) %>% 
+  arrange(libID) %>% 
+  pivot_wider(names_from = libID) %>% 
+  column_to_rownames("name")
+
+meta1 <- dat2 %>% 
+  filter(experiment == "P259_1") %>% 
+  arrange(libID) %>% 
+  droplevels()
+
+count2 <- dat2 %>% 
+  filter(experiment == "P259_2") %>% 
+  select(libID, RhinoVirusA_normCount) %>% 
+  pivot_longer(-libID) %>% 
+  arrange(libID) %>% 
+  pivot_wider(names_from = libID) %>% 
+  column_to_rownames("name")
+
+meta2 <- dat2 %>% 
+  filter(experiment == "P259_2") %>% 
+  arrange(libID) %>% 
+  droplevels()
+
+#### limma ####
+#Run in 3.P259.pDC_response.to.review
+pval_1.contrast <- read_csv('review_response/RV.EOS.model.csv')
+pval_2.contrast <- read_csv('review_response/RV.antiIL5.model.csv')
+
+#### Plot ####
+#Pvals for eos experiment
+GOI.p1 <- pval_1.contrast %>% 
+  mutate(dataset="P259.1")
+
+#Pvals for aniti-IL5 experiment
[…]

### 2022-03-18 publication/scripts/FigE3.STRING.R

@@ -1,57 +1,0 @@
-library(tidyverse)
-library(biomaRt)
-source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/STRING_network_fxn.R")
-set.seed(8434)
-
-#### DEG ####
-fdr.cut <- 0.1
-#Define change with virus
-DEG1.v <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
-  filter(group %in% c("none_HRV - none_none", "EOS.supp_HRV - EOS.supp_none") &
-           adj.P.Val <= fdr.cut)
-#Define change with treament
-DEG1.t <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
-  filter(group %in% c("EOS.supp_none - none_none", "EOS.supp_HRV - none_HRV") &
-           adj.P.Val <= fdr.cut)
-
-#Define change with virus
-DEG2.v <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
-  filter(group %in% c("none_HRV - none_none", "AntiIL5_HRV - AntiIL5_none") &
-           adj.P.Val <= fdr.cut)
-#Define change with treament
-DEG2.t <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
-  filter(group %in% c("AntiIL5_none - none_none", "AntiIL5_HRV - none_HRV") &
-           adj.P.Val <= fdr.cut)
-
-#Format in list
-DEG <- c(intersect(DEG1.v$hgnc_symbol, DEG1.t$hgnc_symbol),
-         intersect(DEG2.v$hgnc_symbol, DEG2.t$hgnc_symbol))
-##replace HGNC symbol with one in STRING
-DEG <- gsub("MARCHF2","MARCH2",DEG)
-DEG <- gsub("H2BC21","HIST2H2BE",DEG)
-DEG <- gsub("H4C14","HIST2H4A",DEG)
-
-#### GSEA terms ####
-H <- read_csv("results/enrichment/enrich_DEG_H.csv") %>% 
-  filter(Description %in% c("HALLMARK_INTERFERON_ALPHA_RESPONSE",
-                            "HALLMARK_INTERFERON_GAMMA_RESPONSE",
-                            "HALLMARK_INFLAMMATORY_RESPONSE",
-                            "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"))  %>% 
-  mutate(Description = recode_factor(factor(Description), 
[…]

### 2022-03-18 publication/scripts/FigE4.GSEA.term.venn.R

@@ -1,0 +1,40 @@
+#### Setup ####
+library(plyr)
+library(tidyverse)
+library(fgsea)
+library(venn)
+
+#### Data ####
+myGO <- fgsea::gmtPathways("data_clean/Broad_gmt/h.all.v7.4.symbols.gmt") %>% 
+  #Convert to data frame
+  plyr::ldply(., data.frame) %>% 
+  dplyr::rename(pathway = `.id`, gene = `X..i..`)
+
+#### Venn ####
+# Get list of genes in pathways
+
+venn.ls <- list()
+
+for(term in c("HALLMARK_INTERFERON_ALPHA_RESPONSE",
+              "HALLMARK_INTERFERON_GAMMA_RESPONSE",
+              "HALLMARK_INFLAMMATORY_RESPONSE",
+              "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION")){
+  
+  #Subset to term of interest
+  myGO.temp <- myGO %>% 
+    filter(pathway == term) %>% 
+    distinct(gene) %>% unlist(use.names = FALSE)
+  
+  #Save to list
+  name <- gsub("HALLMARK_", "", term)
+  name <- gsub("_", "\n", name)
+  
+  venn.ls[[name]] <- myGO.temp
+}
+
+# Plot venn and save
+pdf("publication/fig/FigE4.GSEA.term.venn.pdf",
+    width=5.5, height=5)
+venn(ilab=FALSE, ilcs=1, sncs=1, box = FALSE,
+     x=venn.ls)
+dev.off()

### 2022-03-18 publication/scripts/FigE5.GSEA.fold.change.R

@@ -1,0 +1,308 @@
+#### Setup ####
+library(tidyverse)
+library(cowplot)
+library(limma)
+library(ggrepel)
+set.seed(4389)
+
+#### Data ####
+#List genes in each Hallmark term of interest
+myGO <- fgsea::gmtPathways("data_clean/Broad_gmt/h.all.v7.4.symbols.gmt")
+GO.df <- plyr::ldply(myGO, rbind) %>% 
+  dplyr::rename(term = `.id`) %>% 
+  pivot_longer(-term, values_to="hgnc_symbol")
+
+#Expression data
+load("data_clean/P259_pDC_clean.RData")
+
+#### Format data ####
+#Select genes of interest and format
+## P259.2
+h.combo.plot_2 <- as.data.frame(dat.pDC.voom_2$E) %>% 
+  rownames_to_column("geneName") %>% 
+  #Get HGNC symbols
+  left_join(dat.pDC.voom_2$genes) %>% 
+  dplyr::select(hgnc_symbol, everything()) %>% 
+  #Filter to genes in terms
+  inner_join(GO.df) %>% 
+  dplyr::select(term, hgnc_symbol, geneName, everything(), 
+                -c(`Previous symbols`:gene_biotype), -name) %>% 
+  #longer format
+  pivot_longer(-c(term:geneName), names_to = "libID", 
+               values_to = "expression") %>% 
+  #Add metadata
+  mutate(experiment = "P259.2") %>% 
+  left_join(dplyr::select(dat.pDC.voom_2$targets, donorID, libID, 
+                          IL5, virus.detail))
+
+## P259.1
+h.combo.plot_1 <- as.data.frame(dat.pDC.voom_1$E) %>% 
+  rownames_to_column("geneName") %>% 
+  #Get HGNC symbols
+  left_join(dat.pDC.voom_1$genes) %>% 
+  dplyr::select(hgnc_symbol, everything()) %>% 
+  #Filter to genes in terms
+  inner_join(GO.df) %>% 
+  dplyr::select(term, hgnc_symbol, geneName, everything(), 
+                -c(`Previous symbols`:gene_biotype), -name) %>% 
+  #longer format
+  pivot_longer(-c(term:geneName), names_to = "libID", 
+               values_to = "expression") %>% 
+  #Add metadata
+  mutate(experiment = "P259.1") %>% 
+  left_join(dplyr::select(dat.pDC.voom_1$targets, donorID,
+                          libID, IL5, virus.detail))
+
[…]

### 2022-03-18 publication/scripts/FigE6.STRING.R

@@ -1,0 +1,57 @@
+library(tidyverse)
+library(biomaRt)
+source("https://raw.githubusercontent.com/kdillmcfarland/R_bioinformatic_scripts/master/STRING_network_fxn.R")
+set.seed(8434)
+
+#### DEG ####
+fdr.cut <- 0.1
+#Define change with virus
+DEG1.v <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
+  filter(group %in% c("none_HRV - none_none", "EOS.supp_HRV - EOS.supp_none") &
+           adj.P.Val <= fdr.cut)
+#Define change with treament
+DEG1.t <- read_csv("results/gene_level/P259.1_gene_pval.csv") %>% 
+  filter(group %in% c("EOS.supp_none - none_none", "EOS.supp_HRV - none_HRV") &
+           adj.P.Val <= fdr.cut)
+
+#Define change with virus
+DEG2.v <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
+  filter(group %in% c("none_HRV - none_none", "AntiIL5_HRV - AntiIL5_none") &
+           adj.P.Val <= fdr.cut)
+#Define change with treament
+DEG2.t <- read_csv("results/gene_level/P259.2_gene_pval.csv") %>% 
+  filter(group %in% c("AntiIL5_none - none_none", "AntiIL5_HRV - none_HRV") &
+           adj.P.Val <= fdr.cut)
+
+#Format in list
+DEG <- c(intersect(DEG1.v$hgnc_symbol, DEG1.t$hgnc_symbol),
+         intersect(DEG2.v$hgnc_symbol, DEG2.t$hgnc_symbol))
+##replace HGNC symbol with one in STRING
+DEG <- gsub("MARCHF2","MARCH2",DEG)
+DEG <- gsub("H2BC21","HIST2H2BE",DEG)
+DEG <- gsub("H4C14","HIST2H4A",DEG)
+
+#### GSEA terms ####
+H <- read_csv("results/enrichment/enrich_DEG_H.csv") %>% 
+  filter(Description %in% c("HALLMARK_INTERFERON_ALPHA_RESPONSE",
+                            "HALLMARK_INTERFERON_GAMMA_RESPONSE",
+                            "HALLMARK_INFLAMMATORY_RESPONSE",
+                            "HALLMARK_EPITHELIAL_MESENCHYMAL_TRANSITION"))  %>% 
+  mutate(Description = recode_factor(factor(Description), 
[…]

### 2022-03-18 publication/scripts/Table.formatting.R

@@ -63,7 +63,7 @@ read_csv("results/GSEA/h_GSEA.result.csv") %>%
   #Add experiment variable
   mutate(experiment = ifelse(grepl(".1", contrast), "P259_1", "P259_2")) %>%
   #Set gene set variable
-  mutate(gene.set = "HALLMARK",
+  mutate(gene.set = "H",
          pathway = gsub("HALLMARK_", "", pathway)) %>% 
   #Format contrast to match model results
   mutate(contrast = gsub(".1", "", contrast),
@@ -72,7 +72,20 @@ read_csv("results/GSEA/h_GSEA.result.csv") %>%
          contrast = gsub("[.]A"," - A", contrast),
          contrast = gsub("[.]E"," - E", contrast)) %>% 
   #Reorder variables
-  select(experiment, gene.set, pathway, contrast, everything()) %>% 
-  arrange(experiment, pathway, contrast) %>% 
+  select(experiment, gene.set, pathway, contrast, fgsea.FDR, fgsea.NES, fgsea.leadingEdge,
+         gage.FDR) %>% 
+  arrange(experiment, pathway, contrast, -fgsea.FDR) %>% 
+  filter(fgsea.FDR < 0.1) %>% 
   #Save
   write_csv(., "publication/table/TableE5.GSEA.csv") 
+
+## Hypergeo enrichment
+read_csv("results/enrichment/enrich_DEG_H.csv") %>% 
+  bind_rows(read_csv("results/enrichment/enrich_DEG_C2_CP.csv")) %>% 
+  filter(p.adjust < 0.05 & !is.na(p.adjust)) %>% 
+  rename(gene.set=category, pathway=Description, FDR=p.adjust) %>% 
+  mutate(experiment = ifelse(group == "AntiIL5", "P259_2", "P259_1"),
+         genes = gsub("/",";", SYMBOLs)) %>% 
+  select(experiment, gene.set, pathway, genes, FDR) %>% 
+  #Save
+  write_csv(., "publication/table/TableE6.hypergeo.enrich.csv") 

### 2022-03-18 review_response/Fig.RNA.totalSeqs.R

@@ -1,23 +1,0 @@
-library(tidyverse)
-library(readxl)
-
-#### Data ####
-attach("data_clean/P259_pDC_clean.RData")
-
-dat <- read_excel("data_raw/P259-2 Final Annotation.xlsx") %>% 
-  select("library sampleId", "RNA conc (ng/ul)")
-
-dat <- read_csv("data_clean/P259_pDC_metadata.csv") %>% 
-  select(libID, total_sequences) %>% 
-  inner_join(dat, by=c("libID"="library sampleId")) %>% 
-  mutate(`Quality filter` = ifelse(libID %in% dat.pDC.voom$targets$libID, "Pass", "Fail"))
-
-#### Plot ####
-
-dat %>% 
-  ggplot(aes(x=`RNA conc (ng/ul)`, y=total_sequences)) +
-  geom_point(aes(color=`Quality filter`), size=2) +
-  theme_classic() +
-  labs(y = "Raw sequences", x="RNA concentration (ng/ul)") +
-  ggsave(filename = "review_response/Fig.RNA.totalSeqs.png",
-         height=4, width=5)

### 2022-03-18 apps/boxplot_app.R
Add boxplot app

@@ -1,0 +1,159 @@
+#P259
+library(BiocManager)
+options(repos = BiocManager::repositories())
+#data
+library(readr)
+library(dplyr)
+library(tibble)
+library(tidyr)
+library(limma)
+#plots
+library(ggplot2)
+#shiny
+library(shiny)
+
+#### Data prep ####
+# # Expression data
+# attach("P259_pDC_clean.RData")
+# dat1 <- as.data.frame(dat.pDC.voom_1$E) %>%
+#   rownames_to_column("geneName") %>%
+#   pivot_longer(-geneName, names_to = "libID") %>%
+#   left_join(dat.pDC.voom_1$targets) %>%
+#   left_join(dat.pDC.voom_1$genes) %>%
+#   mutate(experiment="EOS supernatant") %>%
+#   select(experiment, cellType, libID, geneName,
+#          hgnc_symbol, IL5, virus, value) %>%
+#   rename(treatment=IL5)
+# dat2 <- as.data.frame(dat.pDC.voom_2$E) %>%
+#   rownames_to_column("geneName") %>%
+#   pivot_longer(-geneName, names_to = "libID") %>%
+#   left_join(dat.pDC.voom_2$targets) %>%
+#   left_join(dat.pDC.voom_2$genes) %>%
+#   mutate(experiment = "anti-IL5/Ra") %>%
+#   select(experiment, cellType, libID, geneName,
+#          hgnc_symbol, IL5, virus, value) %>%
+#   rename(treatment=IL5)
+# 
+# attach("P259_EOS_clean.RData")
+# dat3 <- as.data.frame(dat.EOS.voom$E) %>%
+#   rownames_to_column("geneName") %>%
+#   pivot_longer(-geneName, names_to = "libID") %>%
+#   left_join(dat.EOS.voom$targets) %>%
+#   left_join(dat.EOS.voom$genes) %>%
+#   mutate(experiment = "EOS supernatant") %>%
+#   select(experiment, cellType, libID, geneName,
+#          hgnc_symbol, IFNa, virus, value) %>%
+#   rename(treatment=IFNa)
+# 
+# dat.all <- bind_rows(dat1,dat2,dat3) %>%
+#   mutate(gene = paste0(hgnc_symbol, " (", geneName, ")"),
+#          virus = recode(virus, "HRV"="RV"),
+#          x=paste(virus, treatment, sep="\n"),
+#          x = factor(x, c("none\nnone",
+#                          "none\nAntiIL5",
[…]
